Beyond Acoustic Cues: Olfactory-Mediated Avoidance of Bats by Crickets

Version 1

Abstract

The evolutionary arms race between insectivorous bats and their insect prey is a classic paradigm of acoustic predation and evasion, with insects having evolved sophisticated auditory countermeasures. Both bats and insects also rely heavily on olfaction for key behaviors, such as social communication. Moreover, predator-derived odors are well-established as risk cues in many other predator–prey systems. However, whether olfaction plays a role in the bat–insect arms race remains unknown. Here, we unveil a previously unknown olfactory dimension to this interaction. We demonstrated that the body odor of the insectivorous bat Scotophilus kuhlii triggered robust avoidance and electrophysiological antennal responses in a common cricket prey, Loxoblemmus equestris. We identified (–)-limonene as a behaviorally active volatile in bat odor that elicited electrophysiological responses in cricket antennae and was sufficient to elicit avoidance in crickets. Field experiments confirmed that (–)-limonene exposure reduced cricket calling activity, demonstrating the ecological relevance of this cue. Our findings establish that insects can detect and initiate avoidance of phylogenetically distant vertebrate predators via olfaction, a process that can be mediated by the elemental perception of individual odor compounds. This work broadens the sensory framework of a classic predator–prey system, and highlights olfactory eavesdropping as a functional strategy in phylogenetically distant predator–prey systems.

Introduction

The evolutionary arms race between insectivorous bats and their insect prey represents a quintessential predator–prey system, characterized by the evolution of sophisticated acoustic countermeasures in insects (Conner and Corcoran, 2012; Ter Hofstede and Ratcliffe, 2016). As dominant nocturnal aerial insectivores, bats consume diverse taxa such as Lepidoptera, Coleoptera, and Diptera (Gong et al., 2025; Srilopan et al., 2025), providing vital biocontrol against agricultural pests like Spodoptera frugiperda and Helicoverpa armigera (Liu et al., 2024; Maine and Boyles, 2015). Over 50 million years of coevolutionary arms race, insects evolved sophisticated auditory countermeasures: ultrasound hearing, sound-absorbing scales, acoustic decoys, and ultrasonic jamming (Barber et al., 2022; Jacobs and Bastian, 2016; Zeng et al., 2011). Auditory detection, independently evolved in at least seven insect orders, is the most widespread adaptation (Ter Hofstede and Ratcliffe, 2016). Upon detecting bat echolocation calls, insects typically execute evasive flight maneuvers away from the predator source to reduce detection and predation risk (Jacobs and Bastian, 2016). However, the dominance of this acoustic narrative may have overshadowed other sensory channels (Stidsholt et al., 2025).

Notably, both bats and insects themselves rely heavily on olfaction for essential behaviors. Many bats (e.g., Hipposideros armiger, Lasiurus cinereus, Phyllostomus hastatus, and Saccopteryx bilineata) produce species-specific volatile organic compounds (VOCs) from specialized glands for social communication (Adams et al., 2018; Caspers et al., 2009; Stefaniak et al., 2025; Zhang et al., 2022). Conversely, insects possess exquisitely sensitive olfactory systems capable of detecting chemical cues (e.g., sex pheromones) over kilometers, critical for foraging, mate finding, and predator avoidance (Kannan et al., 2022). Thus, olfaction could, in principle, provide an additional sensory channel in the bat–insect dynamic. More broadly, predator-derived VOCs are well-established as risk cues that elicit adaptive avoidance behaviors in diverse predator–prey systems (Apfelbach et al., 2005, 2015; Ferrero et al., 2011; Rosen et al., 2015; Takahashi et al., 2005, 2008). Empirical studies demonstrate that prey organisms exhibit multifaceted antipredator responses, including enhanced vigilance, activity suppression, and abandonment of non-defensive behaviors (e.g., foraging), when exposed to predator odors (Cornelis et al., 2019; Hegab et al., 2015; Pérez-Gómez et al., 2015). Such chemically cued defenses are well-established within phylogenetic groups: vertebrate prey (e.g., rabbits, rodents, and ungulates) avoid mammalian predator odors (e.g., bobcats, foxes, and wolves), while invertebrate prey (e.g., various arthropods including crustaceans and insects) evade arthropod predators (e.g., aquatic insect larvae and spiders) (Kempraj et al., 2020; Monclús et al., 2005; Osada et al., 2014; Swihart et al., 1991; Weiss et al., 2018). These intraphyletic frameworks elucidate core principles of chemical communication and inform biological control strategies (Apfelbach et al., 2005; Dicke and Grostal, 2001).

However, a fundamental gap exists in understanding whether such olfactory eavesdropping can operate across the vast phylogenetic divide separating vertebrate predators and invertebrate prey (Apfelbach et al., 2015; Dicke and Grostal, 2001; Schoeppner and Relyea, 2005). Although olfactory interactions across broad taxonomic boundaries are widespread in nature, such as mosquitoes using host odors to blood-feed, elephants and moths sharing pheromonal components, and aroids chemically mimicking carrion to attract pollinating flies (Kang et al., 2023; Zaremska et al., 2022; Zhao et al., 2022), these interactions are primarily shaped by selective pressures tied to foraging, reproduction, or mutualisms, rather than by predation-related selection. Given the pronounced differences between vertebrates and invertebrates in olfactory receptor architectures, neural processing, and behavioral ecology (Kaupp, 2010; Wang et al., 2024), two pivotal and interconnected questions arise: First, can invertebrates detect and respond to volatile cues from phylogenetically distant vertebrate predators? Second, what is the mechanistic basis of such recognition? Specifically, does it require the holistic perception of complex, multi-component odor blends (configural processing), or can it be triggered by the detection of simple, elemental compounds (Apfelbach et al., 2015; Clifford and Riffell, 2013; Coureaud et al., 2022; Lei and Vickers, 2008)? Resolving these questions is essential to advancing theories of sensory evolution and chemical ecology beyond intraphyletic paradigms.

The bat–insect system presents an ideal model to address these questions. Despite the clear importance of olfaction to both taxa and its established role in predator–prey ecology, whether it plays any functional role in the iconic bat–insect arms race remains unexplored. In the present study, we hypothesized that insects might eavesdrop on bat VOCs as an early warning. Using the Asiatic lesser yellow house bat (Scotophilus kuhlii) and its cricket prey (Loxoblemmus equestris) as a model system, we tested whether bat body odor elicits avoidance in insects and if individual odor components were sufficient to trigger such responses. S. kuhlii (Mammalia: Chiroptera: Vespertilionidae) is widely distributed across Asian tropics and subtropics. In China, this species typically forms dense colonies in palm trees (Livistona chinensis). S. kuhlii preys on a variety of insects, including members of the orders Coleoptera, Orthoptera, and Lepidoptera (Zhu et al., 2012). This species forages on the wing over forests, farmlands, and grasslands, habitats shared with L. equestris. This spatial overlap creates predation opportunities, as the bats frequently hunt near the ground. In our study area, L. equestris dominates cricket assemblages in the foraging habitats of S. kuhlii, producing intense mating calls (primarily for mate attraction and territory defense) and frequently relocating by flight, behaviors that heighten its vulnerability to predators (Ren et al., 2023; Stevenson and Rillich, 2012). Notably, S. kuhlii emits a strong and distinctive odor. Following confirmation of this specific predator–prey relationship by DNA metabarcoding, we integrated behavioral, electrophysiological, chemical, and field evidence to demonstrate olfactory-mediated avoidance in L. equestris and identify its underlying mechanism.

Results

L. equestris is a prey species of S. kuhlii

To establish the predatory relationship between S. kuhlii and L. equestris, we conducted DNA metabarcoding analysis on fecal samples from 30 individual bats. Using mitochondrial 16S rRNA gene (16S) and cytochrome c oxidase subunit I gene (COI) markers, we obtained 426,832 insect sequence reads (323 operational taxonomic units, OTUs) and 1,805,577 insect sequence reads (534 OTUs), respectively. The diet of S. kuhlii was primarily composed of Coleoptera, Blattodea, Lepidoptera, Hemiptera, Orthoptera, Mantodea, Diptera, and Hymenoptera (Supplementary Figure 1A). Within Orthoptera, Gryllidae was the dominant family, comprising 94% (16S) and 93.5% (COI) of Orthoptera sequences (Supplementary Figure 1B), and was detected in 50% of bats (15/30 individuals). Phylogenetic analysis revealed at least seven cricket species among the prey (Supplementary Figure 1C). Among these prey crickets, L. equestris was the most abundant species encountered at ground level in the bat foraging sites. This ground-level abundance was supported by insect surveys conducted over eight nights during fecal sample collection, which captured 80 crickets, 70 (87.5%) of which were L. equestris (Supplementary Table 1). Based on the crickets’ well-developed olfactory capacity, the confirmed predation by S. kuhlii, the high local abundance and accessibility of L. equestris, and its tractability in behavioral settings, we selected L. equestris for subsequent experiments testing olfactory responses to bat body odor.

S. kuhlii body odor elicited behavioral avoidance and antennal responses in L. equestris

Behaviorally, in a two-choice olfactometer (Figure 1A), L. equestris strongly avoided the arm containing bat body odor. Upon release, 44 of 47 crickets (93.6%) chose the odor-free control arm, whereas only one entered the bat-odor arm and two shuttled without making a definitive choice (Chi-square test comparing avoiding vs. non-avoiding individuals: χ² = 35.76, df = 1, P < 0.001, φ = 0.87, 95% CI: 0.73 to 1.00; Figure 1B). Control trials with odor-free air in both arms showed no significant bias (13 of 24 vs. 11 of 24 crickets; Chi-square test, χ² = 0.17, df = 1, P = 0.68, φ = 0.08, 95% CI: 0 to 0.49; Figure 1B). Electrophysiological recordings using gas chromatography-electroantennographic detection (GC–EAD) confirmed that volatiles from S. kuhlii body odor are detected by L. equestris antennae. A mixture of bat odor volatiles from eight individuals elicited consistent antennal depolarizations (>0.1 mV) in all five crickets tested (Figure 1C). Parallel analysis by gas chromatography–mass spectrometry (GC–MS) identified six candidate compounds in the same odor samples. Among these, two compounds, namely 2,2-dimethylheptane and (–)-limonene, consistently co-eluted with EAD-active peaks (Figures 1C and 1D), identifying them as key antennal stimulants. No antennal responses were observed in odor-free controls tested on three crickets (Figure 1C). Together, these results establish that L. equestris perceives and avoids the body odor of its predator, S. kuhlii.

Behavioral and electrophysiological responses of L. equestris to S. kuhlii body odor.

(A) Schematic of the Y-tube olfactometer used in two-choice assays. The cricket’s release position (junction) and airflow direction are indicated. (B) Choice percentages of crickets for the control versus test arm across three treatments: bat body odor, air without bat odor (control for odor bias), and (–)-limonene. (C) Representative gas chromatography–electroantennographic detection (GC–EAD) recordings. Upper panel: flame ionization detection (FID) chromatograms of bat body odor extract (brown) and an odor-free control (blue). Lower panel: corresponding EAD response (black) of a L. equestris antenna. Peaks 1–6 correspond to compounds identified by GC–MS (see panel D); asterisks mark compounds that elicited consistent antennal depolarizations (>0.1 mV). (D) Gas chromatography–mass spectrometry (GC–MS) total ion chromatograms of the same bat body odor sample (upper) and an odor-free control (lower). Identified compounds are labeled (peaks 1–6). Axes indicate retention time (x) and relative ion intensity (y).

Snout secretions are the primary source of body odor in S. kuhlii

To identify the biological sources of bat body odor, we analyzed VOCs from hair, feces, and snout (pararhinal gland) secretions of nine bats using headspace solid–phase microextraction coupled with gas chromatography–mass spectrometry (HS–SPME–GC–MS). Snout secretions and hair shared similar hydrocarbon-rich VOC profiles, while fecal volatiles were distinct and lacked (–)-limonene (Figure 2A). Principal component analysis revealed that the VOC profile of a pooled bat body odor sample (from eight bats) clustered more closely with those from snout secretions and hair than with feces (Figure 2B). As hair likely functions as a passive carrier, its surface VOCs probably originate from transferred snout secretions. These results point to snout secretions as the primary source of S. kuhlii body odor.

Volatile organic compound (VOC) profiles of potential odor sources in S. kuhlii.

(A) Hierarchical clustering of VOC profiles from feces, hair, and snout secretions, based on compositional similarity (values normalized 0–1). All three odor sources were sampled from the same nine bats (27 samples total). (B) Principal component analysis (PCA) of VOC profiles from the same samples, based on compound presence/absence data, with the addition of a pooled body odor composite (combined from eight bats) for comparison. Ellipses represent 90% confidence intervals, and marginal density plots are shown for each source.

A single compound, (–)-limonene, triggers avoidance and reduces calling activity in L. equestris

To assess the behavioral relevance of identified volatiles, we measured antennal responses of L. equestris to four commercially available compound previously detected in bat body odor: (–)-limonene, undecane, pentadecane, and hexadecane, using electroantennography (EAG). Compared to a hexane control, only (–)-limonene (10% v/v, 5.87 × 10-7 mol/µL) evoked a significant antennal response (repeated-measures ANOVA, F(4, 28) = 89.77, P < 0.001, η²p = 0.93; (–)-limonene vs. control: Bonferroni-corrected paired t-test, t(7) = 10.70, P < 0.001, Hedges’ g = 3.41; Figure 3A). EAG responses to (–)-limonene were concentration-dependent (repeated-measures ANOVA, F(5, 25) = 24.95, P < 0.001, η²p = 0.83; Figure 3B), with both 1% and 10% (–)-limonene solutions eliciting significantly stronger responses than the hexane control (Bonferroni-corrected paired t-tests, 1%: t(5) = 10.77, P < 0.001, Hedges’ g = 3.82; 10%: t(5) = 6.92, P = 0.005, Hedges’ g = 2.46).

Electrophysiological and ecological responses of L. equestris to candidate bat body odor compounds.

(A) Electroantennographic (EAG) responses to four candidate compounds (n = 8 individuals). Significance versus the hexane control is indicated (**P < 0.001; repeated-measures ANOVA with Bonferroni-corrected paired t-tests). (B) Concentration-dependent EAG responses to (–)-limonene (n = 6 individuals). **P < 0.01, ***P < 0.001 versus control (same statistical test). (C) Mean call rate over time (calls per minute) in field plots treated with (–)-limonene (n = 8 datasets) or a hexane control (n = 7 datasets). Shaded areas represent 95% confidence intervals (1,200 recordings total).

Notably, (–)-limonene alone was sufficient to elicit robust avoidance behavior in L. equestris. In a two-choice olfactometer assay, 34 of 38 L. equestris individuals (89.5%) chose the limonene-free control arm, demonstrating significant avoidance (Chi-square test, χ² = 23.68, df = 1, P < 0.001, φ = 0.79, 95% CI: 0.59 to 0.99; Figure 1B). Field experiments confirmed the ecological relevance of this avoidance. Prior to spray application during their peak nocturnal activity period, the mean call rate of L. equestris did not differ significantly between treatment and control plots (mean ± SD, 245 ± 143 vs. 207 ± 133 calls/min; GLMM, estimate = 0.10, SE = 0.25, 95% CI: - 0.39 to 0.59, Z = 0.38, P = 0.720; Supplementary Table 2). When exposed to (–)-limonene, the mean call rate of L. equestris in treatment plots significantly decreased over the 40-min trial (post-vs. pre-exposure: mean ± SD, 130 ± 139 vs. 245 ± 143 calls/min; GLMM, estimate = -0.68, SE = 0.06, 95% CI: -0.80 to -0.57, Z = -11.90, P < 0.001; Figure 3C and Supplementary Table 2). In contrast, call rates increased significantly in control plots sprayed with hexane alone (post-vs. pre-exposure: 298 ± 126 vs. 207 ± 133 calls/min; GLMM, estimate = 0.40, SE = 0.06, 95% CI: 0.28 to 0.51, Z = 6.68, P < 0.001; Figure 3C and Supplementary Table 2), a pattern consistent with their natural crepuscular activity peak. Together, these results demonstrate that (–)-limonene, a single component of bat odor, is sufficient to induce avoidance and reduce call activity in L. equestris under natural conditions.

Discussion

Our study provides direct experimental evidence that an insect species can detect and avoid a phylogenetically distant vertebrate predator via olfaction. We demonstrate that the body odor of the insectivorous bat S. kuhlii triggers robust avoidance and electrophysiological responses in the cricket L. equestris, and that a single compound, (–)-limonene, is sufficient to elicit this avoidance in the laboratory and suppress calling in the field. These findings fundamentally expand the sensory narrative of a classic predator–prey system and offer novel insights into the mechanisms of olfactory detection and chemical eavesdropping in interactions across broad phylogenetic divides.

Olfaction as a novel and functional channel in bat–insect interactions

For decades, the bat–insect arms race has served as a paradigm for acoustic coevolution, with insect auditory adaptations to echolocation being meticulously documented (Conner and Corcoran, 2012; Ter Hofstede and Ratcliffe, 2016). This dominant narrative, however, may have inadvertently narrowed the research focus, contributing to relatively less exploration of alternative sensory channels beyond audition (Stidsholt et al., 2025). Many bats produce body odor in the form of VOCs from specialized glands for intraspecific communication (Adams et al., 2018; Caspers et al., 2009; Stefaniak et al., 2025; Zhang et al., 2022). Here, we found that the body odor of S. kuhlii, which originates primarily from snout secretions, elicits robust avoidance and antennal responses in the cricket L. equestris. This suggests that insects may eavesdrop on predators’ intraspecific olfactory signals to mitigate predation threat. Our discovery thus reveals a previously overlooked olfactory dimension to this classic predator–prey interaction. This finding aligns with the established importance of multimodal predator detection, challenging the presumed supremacy of acoustics in this particular system (Jacobs and Bastian, 2016).

While our laboratory assays demonstrate clear olfactory-mediated avoidance in crickets, the ecological dynamics of a direct bat–insect encounter in the wild are more complex. Nonetheless, our field experiment provides critical, ecologically relevant support for the anti-predator function of this cue. The significant reduction in cricket calling activity upon limonene exposure—whether due to dispersal from the limonene-treated plots or direct cessation of signaling—constitutes a functionally adaptive anti-predator response, as both outcomes would reduce acoustic attraction and lower predation risk (Belwood and Morris, 1987; Hedrick and Kortet, 2006). This behavioral response confirms the ecological relevance of the olfactory cue, consistent with the evolutionary logic of eavesdropping on predator signals (Hermann and Thaler, 2014).

The functional advantages of this olfactory channel likely complement, rather than replace, established auditory defenses. The high sensitivity and long-range nature of insect olfaction may facilitate earlier predator detection, particularly against bats employing low-intensity or ‘stealth’ echolocation strategies that challenge acoustic detection alone (Corcoran and Conner, 2017; Goerlitz et al., 2010). More broadly, it could provide a vital, generalized defense for insect taxa lacking specialized ultrasonic hearing, including many beetles and non-hearing moths that are common bat prey (Göpfert and Hennig, 2016; Lin et al., 2023). The suppression of calling activity we observed exemplifies a key benefit: reducing acoustic conspicuousness. Insect calls can serve as a potential cue for certain bats, either through passive hearing or by enhancing the prey’s detectability within the returning echo (Alem et al., 2011; Falk et al., 2015; Prakash et al., 2021). We propose that olfaction in this system acts as a general alert mechanism, priming vigilance and enabling more informed deployment of evasive maneuvers triggered by closer-range acoustic (including echolocation) cues. In this integrated model, olfactory and auditory channels might operate across complementary spatial and temporal scales, forming a more robust, layered sensory defense system. Ultimately, direct observations of bat–insect encounters within manipulated olfactory environments will be crucial to fully quantify the survival benefit conferred by this eavesdropping channel under real predation pressure.

Implications for cross-phylum sensory ecology and the mechanism of eavesdropping

Beyond the specific bat–insect model, our work addresses a central question in sensory ecology: how chemical eavesdropping operates within predator–prey systems between phylogenetically distant taxa with fundamentally divergent olfactory systems (Adams et al., 2020; Emerson and Johnson, 2024; Kaupp, 2010). While intraphyletic kairomone detection is well-established (e.g., rodents avoiding carnivore odors, aphids responding to ladybug chemicals), compelling experimental evidence for such olfaction-mediated recognition across broad phylogenetic divides has been limited (Apfelbach et al., 2005; Ferrari et al., 2007; Tanis et al., 2018). By rigorously linking physiological detection to behavioral avoidance within a confirmed predator–prey relationship, we demonstrate that functional cross-phylum olfactory eavesdropping is attainable, proving this capacity persists despite profound divergence in underlying olfactory circuitry.

A key question concerns the perceptual mechanism underlying such olfactory eavesdropping: does it require a complex, species-specific odor blend (“configural processing”), or can it be initiated by simple, “elemental” compounds (Apfelbach et al., 2015; Coureaud et al., 2022; Lei and Vickers, 2008)? Our identification of (–)-limonene as a single, sufficient elicitor of avoidance strongly supports an elemental perception strategy in this cross-phylum context. This mechanism is evolutionarily parsimonious. Evolving high-sensitivity receptors for a reliable “keystone kairomone” bypasses the need to decode complex vertebrate odor profiles (Schoeppner and Relyea, 2005), likely being more efficient than evolving sophisticated neural circuits for configural processing (Haverkamp et al., 2018; Niven and Laughlin, 2008). Functionally, elemental perception enables a rapid, high-sensitivity response to trace amounts (Stengl, 2010), potentially allowing prey to identify threats at lower concentrations or greater distances than required for complex blends (Cardé and Willis, 2008; Minor and Kaissling, 2003). Thus, elemental perception emerges as a plausible and efficient strategy for initiating anti-predator behavior against phylogenetically remote threats.

Our findings establish the sufficiency of (–)-limonene as an avoidance trigger, but do not preclude a role for other odor components. The crickets’ antennal responses to other bat volatiles in our GC–EAD analyses suggest more complex peripheral perception. Additional compounds, either alone or in synergistic blends, may modulate the full behavioral response in nature. Therefore, elemental perception via (–)-limonene likely represents one effective strategy within a broader olfactory toolkit available to insects. This raises key questions for future research: Do blends containing (–)-limonene and other compounds elicit stronger or qualitatively different insect avoidance? Is sensitivity to (–)-limonene an adaptation to bat predation, or an exaptation of a pathway tuned to plant volatiles (Boncan et al., 2020; Schuman, 2023)? We hypothesize that crickets may discriminate bat-derived limonene through context-specific cues (e.g., temporal and spatial patterning, co-occurrence with other bat-specific compounds) to minimize false alarms. Comparative studies on enantiomeric specificity and detection thresholds of cricket olfactory sensory neurons will be essential. Critically, our field data confirm that limonene exposure in nature robustly triggers an adaptive anti-predator response, irrespective of the precise discrimination mechanism.

In conclusion, our integrated approach reveals that the body odor of S. kuhlii triggers avoidance in L. equestris and that a single component, (–)-limonene, is sufficient to elicit this behavior in the laboratory and suppress calling in the field. We have uncovered a novel olfactory axis in the classic bat–insect arms race, demonstrating that chemical cues constitute a functional sensory channel for predator detection. More broadly, it provides compelling evidence that olfactory eavesdropping can operate across vast phylogenetic distances within predator–prey systems. Elemental perception of individual odor compounds may be a parsimonious strategy for such cross-phylum threat detection. Furthermore, identifying behaviorally active volatiles like (–)-limonene provides a foundational clue for the bioinspired development of novel insect repellents (Rasmann et al., 2005). Future work on the neural circuitry, phylogenetic distribution, and ecological contingencies of this capability will elucidate how such olfactory-mediated eavesdropping evolves and operates across the tree of life.

Materials and methods

Diet composition and prey availability of Scotophilus kuhlii

To confirm the predator–prey relationship between S. kuhlii and L. equestris, we analyzed the diet of S. kuhlii via DNA metabarcoding of fecal samples. On July 10, 2022, we captured 30 individuals of S. kuhlii (21 females and nine males) from a roosting colony in a Chinese fan palm in Guangzhou, China, using hand nets. Sampling occurred during the bats’ initial return to the roost following nightly foraging activity, a period associated with high insect consumption. Each bat was placed individually in a clean cloth bag, and fecal samples were collected after two hours. The samples were immediately transferred to 2-mL Eppendorf tubes (Corning, Cat# 430659) and stored at –20 °C. Bats were released unharmed after sample collection.

From each fecal sample, a 100-mg subsample was taken for DNA extraction using an Omega Bio-tek DNA extraction kit according to the manufacturer’s instructions. We targeted a 225-bp fragment of the mitochondrial cytochrome c oxidase subunit I (COI) gene with primers LCO-1490 (5′-GGTCAACAAATCATAAAGATATTGG-3′) and ZBJ-ArtR2c (5′-WACTAATCAATTWCCAAATCCTCC-3′), and a 105-bp fragment of the mitochondrial 16S ribosomal DNA with primers Coleop_16Sc (5′-TGCAAAGGTAGCATAATMATTAG-3′) and Coleop_16Sd (5′-TCCATAGGGTCTTCTCGTC-3′) (Lin et al., 2023). Polymerase chain reaction conditions followed published protocols (Lin et al., 2023). Amplified products of all 64 samples (60 from bat feces and four extraction blanks as negative controls) were sequenced on an Illumina MiSeq PE300 platform following the standard operating procedures at Majorbio Bio-Pharm Technology Co., Ltd. Ultimately, using the 16S molecular marker, we obtained insect sequences from 29 bats; no insect sequences were detected for one individual (bat identity: S28). Using the COI molecular marker, we obtained dietary sequences from 30 individual bats.

Dietary analysis was performed following established bioinformatic protocols (Chang et al., 2019; Lin et al., 2023). Briefly, after quality filtering and merging paired-end reads, DNA sequences were clustered into Operational Taxonomic Units (OTUs) at a 97% similarity threshold using Usearch v11.0 (Edgar, 2016). Taxonomic assignment was conducted by comparing representative sequences from each OTU against the GenBank and BOLD databases, based on the following criteria: (a) sequences with >98% identity to a single species were assigned to that species; (b) those matching multiple congeners within the same genus (>98%) were assigned to the genus level; and (c) sequences matching multiple genera within a family or only family-level references were assigned to the family level. No contaminant insect sequences were detected in the negative controls. Only insect sequences were retained for diet analysis. We quantified bat diet composition using the relative read abundance (RRA) and the weighted percentage of occurrence (wPOO) of prey orders, families and species (Deagle et al., 2018). For RRA, total read counts were normalized to 80,000 reads per sample. The wPOO was calculated from occurrence frequency tables standardized to a total occurrence frequency of 100 across all samples.

To achieve species-level resolution within Gryllidae (crickets), we implemented a phylogenetic approach (Ross et al., 2008). Representative sequences of each Gryllidae OTU were aligned with their top five BLAST matches from NCBI, and duplicate sequences were removed. Multiple sequence alignment was performed in MEGA v7.0 (Kumar et al., 2016), and a maximum-likelihood (ML) phylogeny was reconstructed using IQ-TREE v1.6.12 under the best-fit substitution model (16S: TPM3u+F+G4; COI: TIM2+F+I+G4) (Nguyen et al., 2015), with Gampsocleis sinensis sequences as the outgroup. OTUs clustering within established species-level clades were assigned to the corresponding species.

To assess prey availability, we surveyed insects in a grassland foraging site of S. kuhlii approximately one kilometer from its roost. Sampling was carried out on eight nights between July and September 2022 (19:30–21:30) using one ultraviolet light trap (BY201, Beijing Baoyuan Xingye Technology Co., Ltd., China) and one Malaise trap (BY201, Beijing Baoyuan Xingye Technology Co., Ltd., China). Lepidoptera (moths) and Gryllidae were the most abundant taxa collected. Preliminary Y-tube olfactometer trials indicated that moths exhibited minimal movement in the Y-tube, precluding reliable assessment of odor-mediated behavior, whereas crickets showed consistent locomotor activity. We therefore focused on estimating the relative abundance of Gryllidae species. L. equestris was selected for subsequent olfactory experiments due to its high local abundance and confirmed status as a natural prey item of S. kuhlii (see Results).

Behavioral and electrophysiological assays of cricket responses to S. kuhlii body odor.

We employed two-choice tests in a Y-tube olfactometer (Shelai Instrument Platform, China) to assess whether the body odor of S. kuhlii elicits avoidance behavior in the cricket L. equestris. Individuals of L. equestris were collected nightly (19:30–21:30) from the bat’s foraging habitat (see the section above) and acclimated in insect cages with water provided in a temporary laboratory until testing.

The olfactometer consisted of a Y-shaped glass tube (inner diameter: 35 mm; main arm length: 200 mm; side arm length: 200 mm). One side arm was connected in series to a glass chamber housing one of eight adult S. kuhlii individuals (randomly selected for each trial) and a charcoal filter bottle; the other side arm was connected to an empty glass chamber and a charcoal filter bottle. Both charcoal bottles contained activated carbon for air purification. An air pump supplied airflow, regulated to 1 L/min in each arm by flowmeters.

Behavioral trials commenced at 23:00 under complete darkness. Before each trial, the Y-tube was cleaned with ethanol and oven-dried. Odorless sponges (diameter: 35 mm, thickness: 15 mm; Wuhan Hemida Technology Co., Ltd., China) were placed at the end of each arm to prevent the crickets from escaping. After confirming system airtightness, a single cricket was introduced at the junction (center) of the Y-tube, and its behavior was recorded using an infrared video camera (4K Ultra; Sony, China). Preliminary observations confirmed that crickets typically initiated exploration of the apparatus arms within three seconds of being introduced. Therefore, video recording began once a cricket started exploring the two side arms. A cricket was scored as “avoiding” the bat odor if it entered the control arm and remained there continuously for 10 s. It was scored as “not avoiding” the bat odor if it entered the test arm and remained there continuously for 10 s. Individuals that moved back and forth between the two arms without remaining in either arm for at least 10 s were also classified as “not avoiding”. We selected a 10-second duration as the criterion because the crickets were highly mobile, and this period was sufficient for them to fully explore the olfactometer. A total of 47 crickets were tested, with each individual used in a single trial. The left/right position of the bat odor source was randomized between trials to control for side bias. A separate control experiment with 24 crickets was performed under identical conditions, except that no bat was placed in the glass chamber. Behavioral responses were recorded following the same procedure. To minimize observer bias, the video recordings were analyzed by an experimenter who was blinded to the treatment conditions.

We collected the body odor of S. kuhlii (i.e., its volatile organic compounds, VOCs) for chemical characterization and electrophysiological bioassay using gas chromatography–mass spectrometry (GC–MS) and gas chromatograph–electroantennographic detection (GC–EAD) with cricket antennae, respectively. Samples were collected from eight adult individuals (four females and four males) using a dynamic headspace sampling system equipped with Porapak-Q adsorbent columns. We selected Porapak-Q because bat body odor is rich in hydrocarbons, which are non-polar or only weakly polar, and Porapak-Q has a high adsorption capacity for these compounds. The system consisted of a vacuum pump (QC-1S; Beijing Institute of Labour Protection Science, China), a charcoal filter for air purification, a 5 L glass chamber housing the bats, and a Porapak-Q column (glass tube, 4 mm internal diameter, 200 mg of Porapak-Q 80/100 mesh; Supelco, USA) (Wang et al., 2023). Prior to sampling, each adsorbent column was cleaned by rinsing sequentially with 3 mL of methanol and dichloromethane (DCM), followed by thermal conditioning under nitrogen at 180 °C for 30 minutes. During VOC collection, purified air from the pump was passed through the system at a flow rate of 500 mL/min for two hours. Control samples were collected from empty glass chambers following the same procedure. After sampling, VOCs were eluted from each column with 1 mL of DCM, and the extracts were concentrated under a gentle nitrogen stream to a final volume of 1 mL. All samples were stored at –20 °C until subsequent analysis by GC–EAD and GC–MS. The eight S. kuhlii used for odor collection and behavioral assays were captured from a roosting palm tree on 1 October, 2023. Bats were housed socially in cages (50 × 50 × 50 cm) in a temporary laboratory maintained at 28 °C and 60% relative humidity. Mealworm larvae (Tenebrio molitor) and water were provided daily following nightly experiments.

The antennae are a primary sensory organ for detecting both chemical and mechanical cues in crickets. To determine if L. equestris can physiologically perceive bat body odor, we employed a GC–EAD system following established methods (Barbosa-Cornelio et al., 2019; Ma et al., 2025) to test the electrophysiological sensitivity of L. equestris antennae to bat-derived VOCs. The system consisted of an Agilent 7820A gas chromatograph (Agilent Technologies, USA) and EAD equipment (Syntech, Germany). The EAD setup comprised an IDAC 4 signal recording controller, an MP-15 micromanipulator with high-impedance PRG-3 electrodes.

The tips and bases of cricket antennae were trimmed and mounted on a PRG-3 electrode using an MP-15 micromanipulator. After achieving a stable baseline, odor samples were separated on an Agilent 7820A gas chromatograph. A 4 μL aliquot was injected in splitless mode onto an HP-5 column (30 m × 0.32 mm × 0.25 μm). We selected the HP-5 column because its non-polar stationary phase achieves excellent chromatographic separation of the non-polar to weakly polar hydrocarbons that dominate bat body odor, which is essential for subsequent detection. The GC oven temperature was programmed as follows: held at 50 °C for 5 min, increased at 10 °C/min to 200 °C (held 1 min), then raised at 20 °C/min to 250 °C (held 5 min). Nitrogen carrier gas flow was 2.0 mL/min. Injector and detector temperatures were 250 °C and 300 °C, respectively. The GC effluent was split at a 1:2 ratio using a Y-shaped splitter. Two-thirds of the effluent was directed to a flame ionization detector (FID) for compound detection, and one-third was transferred through a heated transfer line into a humidified airstream (400 mL/min) before being delivered to the cricket antenna. Antennal signals were acquired using an IDAC-4 interface and GcEad software (v. 1.2.5; Syntech, Germany), allowing simultaneous recording of FID and EAD responses. Each chromatographic peak in the FID trace had a corresponding EAD trace showing the depolarization of the antennal potential.

A total of eight antennae from individual crickets were tested. Five antennae were exposed to bat odor extracts, and three were exposed to empty-chamber controls (one antenna per injection). A compound was considered EAD-active if its GC peak coincided with an antennal depolarization >0.1 mV, and this response was reproducible in at least three individual antennae. Active compounds were identified by aligning the retention times and peak profiles of EAD-active signals with those from parallel GC–MS analyses.

Chemical profiling and source analysis of S. kuhlii body odor

To characterize the major volatile constituents of S. kuhlii body odor, we analyzed the collected VOC extracts by GC–MS using an Agilent 6850-5975 mass spectrometer operated in electron ionization (EI) mode at 70 eV. The GC was equipped with the same HP-5 column used for GC–EAD, with helium as the carrier gas at a constant flow of 1 mL/min. A 1 μL aliquot of each sample was injected in splitless mode (0.75-min splitless period) at an injector temperature of 250 °C. The oven temperature program matched that used for GC–EAD. Compounds were tentatively identified by comparison of their mass spectra with the NIST 2017 library (Supplementary Table 3; Supplementary Figure 2–7). The identities of these compounds were confirmed by matching both retention times and mass spectra to those of authentic standards, including injections of a mixture containing 100 ng of each compound.

To investigate potential biological sources of bat body odor, we performed a broadly targeted volatile metabolomic analysis using headspace solid–phase microextraction coupled with GC–MS (HS–SPME–GC–MS) on hair, pararhinal gland (snout) secretions, and feces from nine bats (four females, five males). These bats were collected and maintained as described above. Hair samples (10 mg per individual) were clipped from dorsal and ventral regions. Snout secretions were collected using sterile cotton swabs (CS15-005, Shenzhen SihuaBo Technology Co., Ltd., China), with two blank swabs as controls. Fecal samples (100 mg per individual) were collected as in the dietary analysis. All samples were stored in 2-ml Eppendorf tubes at −80 °C until analysis.

Samples were thawed on ice, transferred to headspace vials, and spiked with 10 µL of an internal standard (50 µg/mL). Volatiles were extracted by HS–SPME: after incubation at 60 °C for 5 min, a 120 μm DVB/CWR/PDMS fiber was exposed to the headspace for 15 min for adsorption, then thermally desorbed at 250 °C for 5 min in the GC injector. Analysis was performed on an Agilent 7890B 7000D GC–MS system equipped with a DB 5MS capillary column (30 m × 0.25 mm × 0.25 μm), with helium at 1.2 mL/min. The oven temperature program was: 40 °C for 3.5 min, then increased at 10 °C /min to 100 °C, 7 °C/min to 180 °C, and 25 °C/min to 280 °C (hold 5 min). The MS operated in EI mode at 70 eV, with ion source, quadrupole, and interface temperatures set at 230 °C, 150 °C, and 280 °C, respectively. Data were acquired in selected ion monitoring (SIM) mode. Each compound was identified by matching the retention time and the presence of one quantitative and two to three qualitative ions to those of a standard reference (Yuan et al., 2022). Data were integrated, corrected, and processed using MassHunter software (B.08.00; Agilent), and metabolites were qualified and quantified against an in-house database (Supplementary Table 4). Retention times for non-target compounds are not fully disclosed due to the proprietary policy of the commercial analytical service provider.

To identify the biological source of the characteristic body odor, we compared the VOC profiles from hair, feces, and snout secretions. Volatiles were classified into mutually exclusive chemical classes for comparison; note that “Hydrocarbons” here refers only to aliphatic hydrocarbons, and “Nitrogen compounds” excludes amines. To objectively assess the overall compositional similarity among hair, feces, snout secretions, and body odor, we performed a principal component analysis (PCA) based on binary (presence/absence) VOC data.

Identifying avoidance-eliciting odor compounds

Chemical profiling of S. kuhlii body odor by GC–MS identified six volatile compounds: 2,2-dimethylheptane, (–)-limonene, undecane, 2,6,7-trimethyldecane, pentadecane, and hexadecane, two of which elicited EAD responses (Figure 1C). Because GC–EAD provides a physiological readout of antennal sensitivity but may not fully capture behavioral relevance across different concentrations or odor contexts, we additionally tested antennal responses to commercially available standards of compounds identified by GC–MS. We selected four compounds ((–)-limonene, undecane, pentadecane, and hexadecane) and evaluated antennal responses using electroantennography (EAG) (Supplementary Table 5). Each compound was diluted to 10% (v/v) in hexane for screening, resulting in a final (–)-limonene concentration of 5.87 × 10⁻7 mol/µL. Hexane alone served as the solvent control. Stimuli were prepared by applying 10 μL of a test solution to a filter-paper strip (10 mm × 30 mm), which was then placed inside a glass Pasteur pipette. The pipette tip was positioned in a humidified airstream (1500 mL/min) directed at the antenna through a stainless-steel tube (7 mm inner diameter, 1 cm from the antenna). Cricket antennae were prepared as for GC–EAD. Each stimulus lasted 0.5 s, with an inter-stimulus interval of 1 min to allow full recovery of antennal responses. For each antenna, the hexane control was presented first, followed by the compounds in sequence. EAG responses were recorded from eight antennae (each from a different cricket) and analyzed using EagPro software (version 2.0; Syntech, Germany).

To examine the behavioral response of L. equestris to (–)-limonene, a two-choice assay was conducted using 10% (v/v) (–)-limonene (5.87 × 10⁻7 mol/µL) as the odor source. A 10 μL aliquot was applied to a filter-paper strip (10 mm × 30 mm) inside a glass chamber. A total of 38 L. equestris individuals were tested under the same experimental conditions as in the bat-odor assay.

To ecologically validate that a single odor compound can elicit avoidance, we tested the effect of 10% (v/v) (–)-limonene (5.87 × 10⁻7 mol/µL) on the calling activity of L. equestris within the bats’ foraging habitat. Over nine nights between 24 June to 16 July 2024, we established four sampling plots (5 m × 5 m) in a linear array at 100-m intervals within the known foraging range of S. kuhlii, at the same location where crickets were originally captured (Supplementary Figure 8). Each night, two plots were randomly assigned to the (–)-limonene treatment and two to hexane control. To assess avoidance, we monitored changes in calling activity of L. equestris as a proxy, using passive acoustic monitoring during their peak nocturnal activity. At each plot, an Audiomoth (v 1.2.0, Open Acoustic Devices, UK) was positioned 0.5 m above ground with the microphone facing downward. Audio was recorded in 55-s files separated by 5-s intervals (1-min duty cycle) at a 384 kHz sampling rate (medium gain). A 40-min pre-exposure recording started at 19:50. Between 20:30 and 20:40, 1 mL of 10% (–)-limonene (treatment) or hexane (control) was evenly sprayed over the 25 m2 area centered on the recorder. A 40-min post-exposure recording started at 20:50. Experimental nights were separated by at least one day, and rainy conditions were avoided.

During acoustic preprocessing, we excluded plot–night datasets that failed quality-control criteria: no cricket calls detected, >20% of 1-min files with zero calls, or total call count <10,000. This quality-control procedure yielded a final sample of 15 datasets: 8 from the (–)-limonene-treated (experimental) plots and 7 from the hexane-treated (control) plots. In total, 1,200 recordings (80 files/plot/night) were analyzed. Using Avisoft SASLab Pro (v 5.2.07, Avisoft Bioacoustics, Germany), recordings were band-pass filtered (3–6.5 kHz) to match the calling frequency of L. equestris (dominant frequency ∼5 kHz, range 4.5–5.5 kHz). This frequency band does not overlap with sympatric cricket species, and L. equestris is the dominant local cricket (see Results), allowing reliable identification. Calls were identified via spectrogram matching using 10 templates of L. equestris calls. A “call” was defined as a train of successive sound elements (syllables). Typical calls of L. equestris last 60–200 ms (Supplementary Figure 9), with inter-syllable intervals < 10 ms within a call, and >60 ms between consecutive calls from the same individual. The number of calls per minute was quantified for each plot and recording period.

Data analysis

All analyses were performed in R v4.3.2 unless specified otherwise. For the two-choice olfactory assay, we used a chi-square goodness-of-fit test to compare the number of crickets exhibiting “avoiding” versus “not avoiding” behavior. Effect sizes are reported as the phi coefficient (φ) with 95% confidence intervals (estimated via the delta method). For chemical profiling, we performed multivariate analysis on VOC data. Hierarchical clustering with heatmap visualization was conducted in Python (v3.9) using the pandas, seaborn, and matplotlib libraries, based on data normalized to a 0–1 range, Euclidean distance, and average linkage. Principal component analysis (PCA) based on a binary (presence/absence) matrix was performed using the vegan package in R to compare profiles from hair, feces, snout secretions, and bat body odor. For the EAG experiment, EAG responses were analyzed using repeated-measures ANOVA (sphericity checked with the rstatix package), followed by Bonferroni-corrected paired t-tests. Effect sizes are reported as partial eta-squared (η²p) for the ANOVA and Hedges’ g for the t-tests. For field calling activity, a generalized linear mixed model (GLMM) was fitted to calls per minute using the glmmTMB package. Model selection compared zero-inflated and standard negative binomial distributions using AIC, and evaluated alternative random effects structures and fixed effects using AICc. The final model included a Group × Phase interaction as a fixed effect, with a random intercept for the plot-by-date combination to account for spatial and temporal non-independence. Model diagnostics were performed with the DHARMa package, including tests for residual uniformity, overdispersion, and zero-inflation.

Animal welfare and ethics

The bats used in this study were carefully handled and maintained under controlled conditions (28 °C, 60% relative humidity) with ad libitum access to food and water. Bats were housed socially in cages (50 cm × 50 cm × 50 cm). No bat mortality occurred during the study, and all bats were released at their original roost sites upon completion of experiments. Crickets were kept at densities ≤ 5 individuals per container to minimize aggression. Crickets not used in EAG assays were returned to their collection sites. All procedures involving live animals were approved by the Science and Technology Ethics Committee of Northeast Normal University, Changchun, China (permit ID: NENU-2022-0308).

eLife Assessment

This valuable study raises the intriguing possibility that crickets use bat-associated odors as cues of predation risk, extending the classic bat-insect arms race beyond its usual acoustic framework. The authors combine fecal metabarcoding, behavioral assays, electrophysiology, chemical analyses, and field observations to show that Loxoblemmus equestris avoids the odor of the insectivorous bat Scotophilus kuhlii, and that synthetic (-)-limonene can elicit antennal responses, avoidance in the laboratory, and reduced calling activity in the field. However, the evidence is currently incomplete because the identity, biological source, natural concentration, and ecological specificity of limonene as a bat-derived predator cue require stronger support, including clearer quantification, contamination controls, individual-level odor data, and evidence that crickets can distinguish bat-associated limonene from common environmental sources. The work will be of interest to researchers in sensory ecology, chemical ecology, predator-prey interactions, and bat-insect coevolution.

Reviewer #1 (Public review):

The manuscript examines whether insects can use bat odor as a cue of predation risk. The authors focus on the insectivorous bat Scotophilus kuhlii and the cricket Loxoblemmus equestris. They first use fecal DNA metabarcoding to show that crickets are part of the bat's diet, and field surveys to show that L. equestris is abundant at local foraging sites. In laboratory Y-tube assays, the authors show that crickets strongly avoid air carrying bat body odor. Gas chromatography coupled with electroantennographic detection showed that cricket antennae respond to components of bat odor. Chemical analyses identified several volatile compounds, with 2,2-dimethylheptane and (−)-limonene associated with antennal responses. Further analyses suggested that snout secretions are likely to contribute to the bat's body odor. The authors then tested individual compounds. Among the commercially available candidates, (−)-limonene elicited a strong antennal response and was sufficient to cause avoidance in the olfactometer. In field plots, spraying (−)-limonene reduced cricket calling activity relative to pre-exposure levels, whereas calling increased in control plots treated with hexane. Overall, the study argues that crickets can detect a vertebrate predator through olfactory cues and that a single bat-associated volatile can trigger antipredator behavior.

This is an interesting and enjoyable study that addresses an understudied aspect of predator-prey interactions. The manuscript is clearly written, the experiments are presented in a logical sequence, and the figures are crisp and easy to follow. I really appreciated the combination of behavioral assays, electrophysiology, chemical analysis, and field observations.

My main issue concerns the identity and biological origin of the proposed bat odor cue, (−)-limonene. Limonene seems like an unusual compound to be emitted endogenously by a mammal, particularly by an insectivorous bat. It would be helpful if the authors could clarify whether mammals are known to synthesize this compound de novo, and, if not, what the likely source of this plant-associated terpene would be in S. kuhlii. Possible sources could include environmental exposure, diet, roosting material, handling, or temporary housing conditions.

I do not doubt that crickets avoid synthetic (−)-limonene. Indeed, this result is quite plausible given that limonene is widely used in insect repellent or repellent-associated fragrance products. However, this also makes contamination an important issue to address explicitly. How did the authors exclude the possibility that limonene entered the samples from human-associated sources, such as insect repellents, soaps, cleaning products, field equipment, cloth bags, cages, gloves, or other materials used while handling wild-caught bats? It would strengthen the manuscript to report limonene levels for individual bat odor collections, all relevant blanks, and any handling or housing controls.

More broadly, given the common occurrence of limonene in plants and human-associated products, I am not yet convinced that it would function as a reliable "keystone kairomone" as suggested around line 253. How would crickets distinguish bat-associated limonene from limonene emitted by a mint leaf, citrus peel, pine material, or other non-threatening environmental sources? The authors may wish to soften this interpretation or provide additional evidence that crickets respond to limonene in a bat-specific context, perhaps through concentration, temporal patterning, co-occurring volatiles, or enantiomeric composition.

Reviewer #2 (Public review):

Summary:

Many insects possess extremely sensitive olfactory systems that can detect chemical signals from distances of several kilometers. For decades, the arms race between bats and insects has served as a prime example of acoustic co-evolution. The auditory adaptations of insects to echolocation have been well documented. Cricket has a multi-sensory predator recognition system with keen olfactory, tactile, and auditory senses. However, whether crickets can use the scent of bats to avoid them remains unknown at present. The authors hypothesized that cricket prey (Loxoblemmus equestris) might eavesdrop on predator bat (Scotophilus kuhlii) VOCs as an early warning. L. equestris is one of the prey species of S. kuhlii, and the authors demonstrated that the body odor of the insectivorous bat S. kuhlii triggers robust avoidance and electrophysiological responses in the cricket L. equestris, and that a single compound, (-)-limonene, is sufficient to elicit this avoidance in the laboratory and suppress calling in the field. Overall, this paper has a complete chain of evidence and should be a highly praised study.

Comments:

(1) Olfactory eavesdropping can transcend the evolutionary divide between vertebrate predators and invertebrate prey, enabling invertebrates to trigger defensive avoidance behaviors in response to predator-derived volatile odors. This phenomenon is empirically well-documented and requires no excessive emphasis.

(2) Without quantitative analysis and without knowing the relative content of this key substance limonene, I don't quite understand how to determine the concentration of limonene standard for EAD, as well as the concentration in field experiments. How is the concentration of limonene determined in field spraying, and is this actually the case in the wild environment?

(3) Figures 1C and D should compare the GC-EAD response of L. equestris to the odor of bat body and the odor of bat nasal secretions. It should not be compared with the air control group. Figure 1D has the same problem.

Author response:

The following is the authors’ response to the original reviews.

We sincerely thank you and the reviewers for the thoughtful evaluation of our manuscript and for the constructive comments and important suggestions. We are encouraged by the recognition that the study is valuable and of interest to researchers in sensory ecology, chemical ecology, predator-prey interactions, and bat-insect coevolution. We are also grateful that the reviewers acknowledged the integrative approach of our work, including fecal metabarcoding, behavioral assays, electrophysiological recordings, chemical analyses, and field observations.

We have carefully considered all comments and have revised the manuscript accordingly. In particular, we have made the following major revisions:

(1) We clarified the biological origin of limonene and expanded the discussion of possible bat-associated sources, including snout secretions and microbial contributions, while avoiding overinterpretation of limonene as an exclusively endogenous mammalian compound.

(2) We added more detailed descriptions of contamination controls, including instrument cleaning procedures, materials used for housing and handling, and blank-control results, to address the concern that limonene could have originated from human-associated or environmental contamination.

(3) We added individual-level odor data and revised the presentation of terpenoid profiles to show more clearly where limonene was detected and how its relative contribution varied among samples.

(4) We clarified the rationale for the concentration choices used in the electrophysiological and field experiments, emphasizing that these assays were designed to test physiological detectability and functional sufficiency rather than to reproduce exact natural emission concentrations or determine response thresholds.

(5) We revised our interpretation of limonene more cautiously. We now state that limonene is sufficient to trigger avoidance responses, while we acknowledge that its natural ecological specificity, concentration dynamics, and interactions with other bat odor components require further investigation.

(6) We corrected the terminology related to limonene enantiomers. Because our GC-MS and GC-EAD analyses did not use an enantioselective column, we replaced “(-)-limonene” with “limonene” throughout the manuscript, figures, legends, and supplementary materials.

(7) We improved the figures and supporting materials by moving the figure illustrating predator-prey relationship into the main text, adding electrophysiological traces from all tested crickets, revising figure legends, clarifying the rationale for comparing bat body odor with air controls, and providing additional chemical-identification details in the supplementary materials.

(8) We checked and clarified statistical annotations, including the exact adjusted P-values for relevant comparisons.

We believe these revisions have substantially improved the clarity, rigor, and balance of the manuscript. We hope that the revised manuscript and the detailed point-by-point responses satisfactorily address all concerns raised.

We are confident that our study represents an important contribution, as it fundamentally expands the traditional acoustic-centred view of bat–insect interactions by demonstrating that crickets can use olfaction as a complementary sensory modality to detect bat odors and initiate avoidance behavior. The multidisciplinary evidence, integrating behavioral assays, electrophysiology, chemical profiling, and field validation, provides convincing support for this novel olfactory pathway.

eLife Assessment

This valuable study raises the intriguing possibility that crickets use bat-associated odors as cues of predation risk, extending the classic bat-insect arms race beyond its usual acoustic framework. The authors combine fecal metabarcoding, behavioral assays, electrophysiology, chemical analyses, and field observations to show that Loxoblemmus equestris avoids the odor of the insectivorous bat Scotophilus kuhlii, and that synthetic (-)-limonene can elicit antennal responses, avoidance in the laboratory, and reduced calling activity in the field. However, the evidence is currently incomplete because the identity, biological source, natural concentration, and ecological specificity of limonene as a bat-derived predator cue require stronger support, including clearer quantification, contamination controls, individual-level odor data, and evidence that crickets can distinguish bat-associated limonene from common environmental sources. The work will be of interest to researchers in sensory ecology, chemical ecology, predator-prey interactions, and bat-insect coevolution.

We sincerely thank the editors for the positive and constructive assessment of our work. We greatly appreciate the recognition of our study’s value and its potential interest to multiple research communities.

We have carefully considered all comments and have revised the manuscript accordingly. The revisions include clarifications of limonene’s biological origin and ecological specificity, strengthened contamination controls and individual-level odor data, clearer rationales for experimental concentration choices, and more cautious interpretation throughout the manuscript. All changes are addressed in detail in our point-by-point responses below.

Importantly, the central conclusion of our study—that crickets can detect and avoid bat odor through olfaction, and that limonene is sufficient to trigger avoidance responses in both laboratory and field settings—remains robustly supported by the multidisciplinary evidence we present. We believe the revised manuscript now provides a clearer, more balanced, and scientifically rigorous account of our findings, and we hope it meets the standards of eLife.

Public Reviews:

Reviewer #1 (Public review):

The manuscript examines whether insects can use bat odor as a cue of predation risk. The authors focus on the insectivorous bat Scotophilus kuhlii and the cricket Loxoblemmus equestris. They first use fecal DNA metabarcoding to show that crickets are part of the bat's diet, and field surveys to show that L. equestris is abundant at local foraging sites. In laboratory Y-tube assays, the authors show that crickets strongly avoid air carrying bat body odor. Gas chromatography coupled with electroantennographic detection showed that cricket antennae respond to components of bat odor. Chemical analyses identified several volatile compounds, with 2,2-dimethylheptane and (−)-limonene associated with antennal responses. Further analyses suggested that snout secretions are likely to contribute to the bat's body odor. The authors then tested individual compounds. Among the commercially available candidates, (−)-limonene elicited a strong antennal response and was sufficient to cause avoidance in the olfactometer. In field plots, spraying (−)-limonene reduced cricket calling activity relative to pre-exposure levels, whereas calling increased in control plots treated with hexane. Overall, the study argues that crickets can detect a vertebrate predator through olfactory cues and that a single bat-associated volatile can trigger antipredator behavior.

This is an interesting and enjoyable study that addresses an understudied aspect of predator-prey interactions. The manuscript is clearly written, the experiments are presented in a logical sequence, and the figures are crisp and easy to follow. I really appreciated the combination of behavioral assays, electrophysiology, chemical analysis, and field observations.

We sincerely thank you for your very positive and encouraging evaluation of our work. We are delighted that you found the study interesting and enjoyable. We also appreciate your kind remarks about the clarity of the manuscript, the logical flow of the experiments, and the quality of the figures.

We are especially grateful that you recognized the value of our integrative approach. Combining behavioral assays, electrophysiology, chemical analysis, and field observations was central to our study design, and we are pleased that this approach resonated with you.

You provided an accurate and comprehensive summary of our work. You confirmed that our main narrative is clear and logically coherent. Specifically, you followed our progression from establishing the predator–prey relationship, to demonstrating olfactory avoidance, to identifying limonene as an active compound, and finally to validating its behavioral effects in both laboratory and field settings.

We have carefully considered all your constructive comments and suggestions. We address them in detail in our point-by-point responses below. Your feedback has been extremely helpful, and we believe the revised manuscript is substantially stronger as a result.

My main issue concerns the identity and biological origin of the proposed bat odor cue, (−)-limonene. Limonene seems like an unusual compound to be emitted endogenously by a mammal, particularly by an insectivorous bat. It would be helpful if the authors could clarify whether mammals are known to synthesize this compound de novo, and, if not, what the likely source of this plant-associated terpene would be in S. kuhlii. Possible sources could include environmental exposure, diet, roosting material, handling, or temporary housing conditions.

I do not doubt that crickets avoid synthetic (−)-limonene. Indeed, this result is quite plausible given that limonene is widely used in insect repellent or repellent-associated fragrance products. However, this also makes contamination an important issue to address explicitly. How did the authors exclude the possibility that limonene entered the samples from human-associated sources, such as insect repellents, soaps, cleaning products, field equipment, cloth bags, cages, gloves, or other materials used while handling wild-caught bats? It would strengthen the manuscript to report limonene levels for individual bat odor collections, all relevant blanks, and any handling or housing controls.

More broadly, given the common occurrence of limonene in plants and human-associated products, I am not yet convinced that it would function as a reliable "keystone kairomone" as suggested around line 253. How would crickets distinguish bat-associated limonene from limonene emitted by a mint leaf, citrus peel, pine material, or other non-threatening environmental sources? The authors may wish to soften this interpretation or provide additional evidence that crickets respond to limonene in a bat-specific context, perhaps through concentration, temporal patterning, co-occurring volatiles, or enantiomeric composition.

We sincerely thank you for your critical and constructive comments. Your questions regarding the identity, biological origin, and ecological specificity of limonene are insightful and have helped us substantially strengthen the manuscript. We address each of your points below.

On the biological origin of limonene and whether mammals synthesize it de novo

You raised an important point that limonene seems unusual for a mammal to emit endogenously. We fully agree. We agree that direct evidence for de novo synthesis of limonene in mammals is currently limited.

However, limonene in bat body odor could still have biological origins. First, it may originate from skin- or gland-associated microbiota. Recent work has shown that skin-associated microorganisms can substantially shape bat volatile odor profiles (Sun et al., 2026, BMC Biology), and some microbes possess enzymes capable of terpene biosynthesis. Second, previous studies have independently reported limonene in the secretions of several bat species (Faulkes et al., 2019, PeerJ; Zhang et al., 2022, Ann. N.Y. Acad. Sci.). This suggests that its presence in bats is not unique to our study. Third, our own analyses detected limonene in hair and snout secretions, but not in faeces or blank controls. This pattern is consistent with a biological source associated with the body surface, rather than diet or environmental deposition.

We have now expanded our Discussion to cover these possibilities more explicitly. We also emphasize that the exact source, i.e., endogenous, microbial, or otherwise, remains an open question that warrants future investigation. Please see lines 258–273 of the clean version of the revised manuscript, or see the excerpt below:

“Although limonene reliably induced avoidance behaviour in crickets, two related questions still merit careful consideration. One question is whether the limonene we identified genuinely originates from bats or reflects contamination during sampling. Limonene is common in plants and numerous consumer products (Boncan et al., 2020; Schuman, 2023), making its endogenous production by a mammal seem unusual. Nevertheless, multiple lines of evidence militate against contamination. First, we adhered to rigorous protocols. For example, all instruments were cleaned with ethanol and oven-dried before each use; bats were housed in stainless-steel cages, and cloth bags had been rinsed with purified water. Second, limonene was absent from all blank controls, including empty-chamber air samples and clean swabs, and was not detected in bat faecal samples. In contrast, it was consistently identified in hair and snout-secretion samples from bats. Third, independent studies have similarly identified limonene in the secretions of other bat species (Faulkes et al., 2019; Zhang et al., 2022). Furthermore, emerging evidence indicates skin-associated microbes may contribute to bat volatile profiles, with some taxa possessing enzymes involved in terpene biosynthesis (Sun et al., 2026). Taken together, these observations point towards an endogenous or microbe-mediated source, although the exact biosynthetic pathway remains to be determined.”

On contamination from human-associated sources

You asked how we excluded the possibility that limonene entered our samples through handling, equipment, cleaning products, or other human-associated sources. We appreciate this concern and have addressed it in detail.

We believe contamination is highly unlikely for several reasons. First, we followed strict protocols throughout. All instruments were cleaned with ethanol and oven-dried before and after each use. We used stainless-steel cages and cloth bags made of degreased bleached cotton washed with purified water. These materials are not sources of terpenes. Second, we ran multiple blanks. Limonene was not detected in any empty-chamber air controls or in blank cotton swabs. In contrast, it was consistently found in multiple bat snout-secretion samples. This clear difference between samples and blanks strongly argues against contamination. Third, we now report individual-level odor data (see new Figure 3; Supplementary Table 5.xlsx). These data show that limonene was consistently present across individual bats. It did not appear sporadically, as one would expect from accidental contamination.

In the revised manuscript, we have added detailed descriptions of our contamination controls in the Methods section (Please see lines 405–407, 443–444, 471–477 of the clean version of the revised manuscript, or see the excerpt below). We have also included the blank-control results (Supplementary Table 5.xlsx), as you suggested.

Lines 405–407: “Prior to sampling, all glassware was thoroughly rinsed with ethanol and dried in an oven at 120°C, and volatile odor collection was conducted in a dedicated odor-free room to minimize environmental contamination.”

Lines 443–444: “The empty-chamber controls were used to account for potential background signals from the experimental system and to provide a baseline for comparison with bat odor extracts.”

Lines 471–477: “Upon capture, bats were placed in clean stainless-steel cages and kept in groups consistent with their natural social associations during the brief interval prior to immediate odor sampling. Hair samples (10 mg per individual) were clipped from dorsal and ventral regions. Snout secretions were collected using sterile cotton swabs (CS15-005, Shenzhen SihuaBo Technology Co., Ltd., China), with two blank swabs as controls. These blank swab controls were included to account for potential volatile contamination from ambient air or the swab material itself (Supplementary Table 5).”

On how crickets distinguish bat-associated limonene from environmental sources

You raised a thoughtful question about ecological specificity. Given that limonene is abundant in mint, citrus peel, pine, and other non-threatening plants, how would crickets use it as a reliable indicator of bat presence?

We agree with you completely. We do not claim that limonene alone serves as an unambiguous bat-specific signal. Instead, our interpretation is more nuanced. We argue that elemental perception represents one effective strategy within a broader olfactory toolkit. It does not exclude the importance of other cues.

In our revised manuscript, we have softened our interpretation accordingly. We now state explicitly that limonene is sufficient to trigger avoidance under our experimental conditions, but we do not interpret it as a uniquely bat-specific keystone kairomone. We also discuss mechanisms that could help crickets reduce false alarms under natural conditions. These include concentration differences, temporal patterning (bats are active at night), spatial context (specific foraging habitats), co-occurrence with other bat-specific volatiles, and possibly enantiomeric composition. Please see lines 274–292 of the clean version of the revised manuscript, or see the excerpt below.

Lines 274–292: “The second question is how crickets might distinguish bat-derived limonene from environmental sources of this compound, given its prevalence in mint, citrus peel, pine and other non-threatening plants (Boncan et al., 2020; Schuman, 2023). It seems implausible that crickets could simply rely on limonene per se to differentiate a bat from a leaf. Two non-exclusive mechanisms could help resolve this issue. First, limonene need not be the only olfactory cue mediating risk perception. Our findings establish the sufficiency of limonene as an avoidance trigger, but do not preclude a role for other odor components. The crickets’ antennal responses to other bat volatiles in our GC–EAD analyses suggest more complex peripheral perception. Additional compounds, either alone or in synergistic blends, may modulate the full behavioral response in nature. Therefore, elemental perception via limonene likely represents one effective strategy within a broader olfactory toolkit available to insects. Second, crickets may discriminate bat-derived limonene through context-specific cues (e.g., temporal and spatial patterning, co-occurrence with other bat-specific compounds) to minimize false alarms. Comparative studies on enantiomeric specificity and detection thresholds of cricket olfactory sensory neurons will be essential. Equally critical will be future efforts to quantify natural bat odor composition, limonene release rates, ambient exposure concentrations, and odor-plume dynamics, which together will inform ecologically valid stimulus design in controlled assays. Critically, our field data confirm that limonene exposure in nature robustly triggers an adaptive anti-predator response, irrespective of the precise discrimination mechanism.”

We acknowledge that fully testing these ideas would require substantial additional work. We have therefore framed this as an important direction for future research, rather than as a resolved issue in the present study.

We thank you again for these insightful comments. Your feedback has helped us present a more rigorous, balanced, and transparent account of our work.

Reviewer #2 (Public review):

Summary:

Many insects possess extremely sensitive olfactory systems that can detect chemical signals from distances of several kilometers. For decades, the arms race between bats and insects has served as a prime example of acoustic co-evolution. The auditory adaptations of insects to echolocation have been well documented. Cricket has a multi-sensory predator recognition system with keen olfactory, tactile, and auditory senses. However, whether crickets can use the scent of bats to avoid them remains unknown at present. The authors hypothesized that cricket prey (Loxoblemmus equestris) might eavesdrop on predator bat (Scotophilus kuhlii) VOCs as an early warning. L. equestris is one of the prey species of S. kuhlii, and the authors demonstrated that the body odor of the insectivorous bat S. kuhlii triggers robust avoidance and electrophysiological responses in the cricket L. equestris, and that a single compound, (-)-limonene, is sufficient to elicit this avoidance in the laboratory and suppress calling in the field. Overall, this paper has a complete chain of evidence and should be a highly praised study.

We sincerely thank you for your very positive and encouraging evaluation of our work. We are especially gratified that you recognized our study as having a "complete chain of evidence" and as a "highly praised study." This recognition means a great deal to us, given the multidisciplinary nature of our approach and the effort required to integrate behavioral, electrophysiological, chemical, and field data into a coherent narrative.

We also appreciate your accurate summary of our work. You correctly highlighted the broader context that while acoustic co-evolution between bats and insects is well documented, whether crickets can use bat scent as an early warning cue has remained unknown. Your summary confirms that our main findings are clear: the body odor of S. kuhlii triggers robust avoidance and electrophysiological responses in L. equestris, and that limonene alone is sufficient to elicit avoidance in the laboratory and suppress calling in the field.

We are particularly grateful that you acknowledged the multi-sensory nature of cricket predator recognition, with keen olfactory, tactile, and auditory senses. We agree that crickets are an excellent model for studying multimodal predator detection, and we hope our study encourages further exploration of olfaction in this classic predator–prey system.

We have carefully considered all your specific comments and suggestions. These include questions about the novelty framing of olfactory eavesdropping, the rationale for our concentration choices, and the presentation of electrophysiological comparisons. We address each of these points in detail in our point-by-point responses below. Your thoughtful feedback has been extremely helpful in improving the clarity and rigor of the manuscript.

Comments:

(1) Olfactory eavesdropping can transcend the evolutionary divide between vertebrate predators and invertebrate prey, enabling invertebrates to trigger defensive avoidance behaviors in response to predator-derived volatile odors. This phenomenon is empirically well-documented and requires no excessive emphasis.

Thank you for this comment. You are absolutely right that olfactory eavesdropping across the vertebrate–invertebrate divide is not a new concept in itself, and we acknowledge that this phenomenon has been well documented in previous studies.

However, we would like to clarify our intended emphasis. In the Introduction and Discussion, we have already stated that empirical examples combining chemical identification, electrophysiological validation, behavioral assays, and field confirmation within a direct predator–prey context remain relatively limited. This is especially true for the bat–insect system, where research has historically focused on acoustic interactions rather than olfaction.

We did not intend to overstate the novelty of olfactory eavesdropping per se. Instead, our emphasis was on providing a complete chain of evidence in a vertebrate–invertebrate predator–prey system that has traditionally been viewed through an acoustic lens. In that sense, we believe our study adds a complementary olfactory perspective to this classic system, rather than claiming to have discovered olfactory eavesdropping as a novel phenomenon. We hope this clarifies our position, as already stated in the original manuscript (lines 71–78, lines 88–90, lines 235–241 of the clean version of the revised manuscript, or see the excerpt below):

Lines 71–78: “However, a fundamental gap exists in understanding whether such olfactory eavesdropping can operate across the vast phylogenetic divide separating vertebrate predators and invertebrate prey (Apfelbach et al., 2015; Dicke and Grostal, 2001; Schoeppner and Relyea, 2005). Although olfactory interactions across broad taxonomic boundaries are widespread in nature, such as mosquitoes using host odors to blood-feed, elephants and moths sharing pheromonal components, and aroids chemically mimicking carrion to attract pollinating flies (Kang et al., 2023; Zaremska et al., 2022; Zhao et al., 2022), these interactions are primarily shaped by selective pressures tied to foraging, reproduction, or mutualisms, rather than by predation-related selection.”

Lines 88–90: “The bat–insect system presents an ideal model to address these questions. Despite the clear importance of olfaction to both taxa and its established role in predator–prey ecology, whether it plays any functional role in the iconic bat–insect arms race remains unexplored.”

Lines 235–241: “Beyond the specific bat–insect model, our work addresses a central question in sensory ecology: how chemical eavesdropping operates within predator–prey systems between phylogenetically distant taxa with fundamentally divergent olfactory systems (Adams et al., 2020; Emerson and Johnson, 2024; Kaupp, 2010). While intraphyletic kairomone detection is well-established (e.g., rodents avoiding carnivore odors, aphids responding to ladybug chemicals), compelling experimental evidence for such olfaction-mediated recognition across broad phylogenetic divides has been limited (Apfelbach et al., 2005; Ferrari et al., 2007; Tanis et al., 2018).”

(2) Without quantitative analysis and without knowing the relative content of this key substance limonene, I don't quite understand how to determine the concentration of limonene standard for EAD, as well as the concentration in field experiments. How is the concentration of limonene determined in field spraying, and is this actually the case in the wild environment?

Thank you for this question. We fully agree that knowing the natural concentrations and relative content of limonene in bat odor would be valuable. However, our experimental aims were not to mimic natural emission levels precisely. Instead, they were designed to answer two distinct questions: First, whether cricket antennae are physiologically capable of detecting limonene at all; and second, whether limonene alone is sufficient to trigger behavioral responses under controlled and semi-natural conditions.

For the EAG experiments, we selected a concentration gradient (0.001%, 0.01%, 0.1%, 1%, and 10% v/v) following standard practices in insect chemical ecology and referencing a previous study (Tang et al., 2024). The goal was to establish dose-dependent antennal sensitivity, not to match a specific natural concentration. Our data clearly show that cricket antennae respond across a range of concentrations, with stronger responses at higher doses.

For the field experiment, we used a 10% v/v limonene spray over 25 m2 plots. We acknowledge that this concentration does not quantitatively reflect natural bat emissions. Natural odor plumes are highly dynamic and depend on airflow, turbulence, temperature, humidity, vegetation structure, and distance from the source. Accurately reconstructing these natural dynamics would require detailed quantitative measurements of bat odor release rates and plume modeling, which were beyond the scope of the present study. Instead, our field experiment was designed for a functional purpose: to test whether limonene could alter cricket calling behavior under semi-natural conditions, using a concentration sufficient to produce a detectable odor stimulus in the field.

We also note that the field-applied concentration is comparable to what has been used in other chemical ecology studies testing the behavioral effects of single volatile compounds under natural or semi-natural conditions. In that context, our positive result supports the ecological relevance of limonene as an avoidance cue, without requiring that the exact applied concentration matches natural bat emissions.

We agree that quantitative characterization of natural bat odor composition, limonene release rates, and ambient exposure concentrations is an important direction for future research. According to your comments, we have added this point to the revised Discussion as a clear future direction (please see lines 286–290 of the clean version of the revised manuscript, or see the excerpt below). We have also clarified in the Methods that our assays were designed to test physiological detectability and functional sufficiency, rather than to establish concentration thresholds or mimic natural emissions exactly (please see lines 515–522, 556–558 of the clean version of the revised manuscript, or see the excerpt below).

Lines 286–290: “Comparative studies on enantiomeric specificity and detection thresholds of cricket olfactory sensory neurons will be essential. Equally critical will be future efforts to quantify natural bat odor composition, limonene release rates, ambient exposure concentrations, and odor-plume dynamics, which together will inform ecologically valid stimulus design in controlled assays.”

Lines 515–522: “For each antenna, a hexane control was first presented to establish baseline antennal activity. For the initial screening, limonene, undecane, pentadecane, and hexadecane were diluted to 10% (v/v) in hexane and delivered individually in a randomized order. To assess dose-dependent responses, limonene was further tested at five concentrations (0.001%, 0.01%, 0.1%, 1%, and 10%, v/v in n-hexane), following the concentration gradient used in a previous study (Tang et al., 2024). Following the initial hexane control, the five limonene concentrations were tested in a randomized order across trials. Each stimulus lasted 0.5 s, with an inter-stimulus interval of 1 min to allow full recovery of antennal responses.”

Lines 556–558: “Our assays were designed to test physiological detectability and functional sufficiency, rather than to establish concentration thresholds or mimic natural emissions exactly.”

(3) Figures 1C and D should compare the GC-EAD response of L. equestris to the odor of bat body and the odor of bat nasal secretions. It should not be compared with the air control group. Figure 1D has the same problem.

Thank you for this suggestion. We understand your point that comparing GC-EAD responses between bat body odor and snout secretions would be a more direct way to identify the anatomical source of active compounds.

However, we would like to explain why we did not include this comparison in the current study.

First, the purpose of Figures 1C and 1D (i.e., Figure 2C and 2D in the revised manuscript) was to answer a more fundamental question: whether bat body odor, as a whole, contains volatile compounds that are detectable by cricket antennae. Comparing with an odor-free air control was therefore the appropriate first step. It established the basic phenomenon of olfactory detection before we moved on to source attribution.

Second, we did conduct chemical profiling of snout secretions, hair, and faeces using HS-SPME-GC-MS (presented in Figure 3). These analyses showed that limonene was consistently present in hair and snout secretions, but absent from faeces and blanks. This allowed us to identify snout secretions as the most likely source of limonene, without requiring GC-EAD recordings from secretion samples themselves.

Third, we did not perform GC-EAD on snout secretions for practical reasons. The secretion samples were collected in very small amounts. They were almost entirely consumed during the HS-SPME-GC-MS chemical analyses, leaving insufficient material for additional GC-EAD testing.

We agree with you that directly comparing GC-EAD responses to snout secretions versus whole-body odor would be an excellent experiment. It would further strengthen the source attribution and provide more direct evidence. We have noted this as a valuable direction for future studies in the revised Discussion.

We hope this clarifies our rationale. Thank you again for your thoughtful suggestion.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) I would suggest moving Supplementary Figure 1 into the main figures. It contains important information about the predator-prey relationship and the ecological relevance of L. equestris, so it seems too central to be placed only in the supplement.

We agree with your suggestion. The predator–prey relationship and the ecological relevance of L. equestris are indeed central to the biological context of this study. We have therefore moved the original Supplementary Figure 1 into the main text as Figure 1. We have renumbered the remaining figures accordingly, revised the figure legends, and updated the Results text to better highlight this ecological context (revised manuscript, Figure 1 legend, lines 856–864).

(2) Lines 134 to 136: Only one representative EAD trace is shown. I suggest showing all five traces, either in the main figure or as a supplementary figure, to better illustrate the reproducibility of the antennal responses across individuals.

We agree. To better illustrate reproducibility across individuals, we have added EAD traces from all five tested crickets as Supplementary Figure 1. The figure legend now describes the sample sizes for both the bat odor treatment and the odor-free control (revised manuscript, Supplementary Figure 1 legend, lines 900–906).

(3) Lines 144 to 153: It would be helpful if the authors reported which VOC collections contained limonene and in what amounts. Showing individual-level data for the bat odor samples, rather than only pooled or summarized profiles, would strengthen the conclusion that limonene is consistently associated with S. kuhlii body odor.

We agree. To better show the consistency of limonene detection across individuals, we have added Figure 3B, which displays an individual-level terpenoid profile. Each stacked bar represents one VOC collection, and the limonene segment indicates its presence and relative contribution. We have also revised the Results section to state explicitly that limonene was detected in hair and snout secretions but absent from feces (lines 144–149 in the revised manuscript).

Lines 144–149: “To identify the biological sources of bat body odor, we analyzed VOCs from hair, faeces, and snout (pararhinal gland) secretions of nine bats using headspace solid–phase microextraction coupled with gas chromatography–mass spectrometry (HS–SPME–GC–MS). Snout secretions and hair shared similar hydrocarbon-rich VOC profiles, whereas faecal volatiles were distinct (Figure 3A). Individual-level terpenoid profiles further showed that limonene was detected in hair and snout secretion VOC collections but was absent from faeces (Figure 3B and Supplementary Table 2).”

(4) Figure 2A: I recommend adding representative chromatograms or VOC traces for feces, hair, and snout secretions. This would make the source comparison more transparent and would help readers assess the underlying chemical profiles behind the heatmap.

We agree that representative chromatograms would make the source comparison more transparent. However, due to the analytical workflow, individual chromatograms were not retained in the dataset we received. As an alternative, we added Figure 3B, which shows individual-level terpenoid profiles. This allows readers to assess which samples contained limonene and how its relative contribution varied among sample types and individuals. In addition, we have provided the NIST retention index, quantitative ion, qualitative ion, and molecular weight for each terpenoid compound in Supplementary Table 2.

(5) The chemical identification of the key compounds would benefit from more detail. The authors state that compound identities were confirmed by matching retention times and mass spectra to authentic standards, including a mixed standard injection. It would be useful to provide the retention times, match and reverse-match scores, blank traces, and, if available, sample-plus-standard co-injection data showing peak augmentation without the appearance of new peaks. For (−)-limonene specifically, the enantiomeric assignment would require an enantioselective method, such as chiral gas chromatography, unless this was already performed and not described.

We fully agree that comprehensive identification evidence is important for transparency and reproducibility.

Regarding the identification data: we have already confirmed compound identities by matching retention times and mass spectra to authentic standards, including a mixed standard injection. In the revised version, we will deposit the raw chromatographic data and identification details (retention times, match scores, and blank traces) in Figshare as supporting information.

Regarding co-injection: we acknowledge that sample-plus-standard co-injection would provide even stronger confirmation. However, our bat odor samples were difficult to obtain and were almost entirely consumed during the GC–EAD and GC–MS analyses. We therefore could not perform additional co-injection validation. We have noted this limitation in the revised Materials and Methods (revised manuscript, lines 463–464).

Regarding the enantiomer assignment of limonene: you are correct that determining the specific enantiomer requires a chiral GC column, which was not available in this study. To avoid overinterpretation, we have replaced “(-)-limonene” with “limonene” throughout the manuscript.

Reviewer #2 (Recommendations for the authors):

(1) This is a typical study in the field of chemical ecology. As long as the source of the active substances is determined and the biological activity has been detected, this research is complete, so Figure 2 seems to be unnecessary.

We agree that identifying the source of the active compound and confirming its biological activity are central to this study. However, we believe Figure 2 serves an important purpose. Bat body odor could originate from multiple sources, i.e., hair, faeces, or snout secretions, and comparing VOC profiles across these sources helps us determine which source most likely contributes to the odor cues detected by crickets. This is especially important for limonene, because it is a plant-associated terpenoid and not a typical animal-derived volatile, as the reviewer #2 pointed out. Our analysis showed that hair and snout secretions shared similar VOC profiles, while faecal VOCs were distinct and lacked limonene. This supports snout secretions as the likely primary source. To make this purpose clearer, we have revised the Methods section to state that this analysis was conducted to investigate potential biological sources of bat body odor (revised manuscript, lines 494–495, 567–569).”

Line 494–495: “To identify the biological source of the characteristic body odor, we compared the VOC profiles from hair, faeces, and snout secretions.”

Line 567–569: “Principal component analysis (PCA) based on a binary (presence/absence) matrix was performed using the vegan package in R to compare profiles from hair, feces, snout secretions, and bat body odor.”

(2) Figure 3B seems to be incorrect. The significance of n-hexane and 1% limonene is ***P < 0.001, while for 10% limonene, why is it only two stars, **P < 0.01? Please check it.

We have rechecked the statistical analysis and confirmed that the original annotation was correct. The significance levels in original Figure 3B (Figure 4B in the revised manuscript) were based on Bonferroni-corrected paired t-tests comparing each limonene concentration with the n-hexane control. The adjusted P value was 0.0006 for 1% limonene (P < 0.001) and 0.00485 for 10% limonene (P < 0.01). The higher adjusted P value for 10% limonene reflects greater among-individual variation at this concentration, which may be due to differential sensitivity of individual antennae at higher doses. We have now added the exact adjusted P values in the revised manuscript, lines 163–167.

Lines 163–167: “EAG responses to limonene were concentration-dependent (repeated-measures ANOVA, F(5, 25) = 24.95, P < 0.001, η2p = 0.83; Figure 4B), with both 1% and 10% limonene solutions eliciting significantly stronger responses than the hexane control (Bonferroni-corrected paired t-tests, 1%: t(5) = 10.77, P < 0.001, Hedges' g = 3.82; 10%: t(5) = 6.92, P = 0.005, Hedges' g = 2.46).”

Finally, we would like to express our sincere gratitude to the editors and the reviewers for the thoughtful feedback. Your comments have significantly improved the quality and clarity of our manuscript. We hope the revised version satisfactorily addresses all concerns raised.

Version 2

Abstract

The evolutionary arms race between insectivorous bats and their insect prey is a classic paradigm of acoustic predation and evasion, with insects having evolved sophisticated auditory countermeasures. Both bats and insects also rely heavily on olfaction for key behaviors, such as social communication. Moreover, predator-derived odors are well-established as risk cues in many other predator–prey systems. However, whether olfaction plays a role in the bat–insect arms race remains unknown. Here, we unveil a previously unknown olfactory dimension to this interaction. We demonstrated that the body odor of the insectivorous bat Scotophilus kuhlii triggered robust avoidance and electrophysiological antennal responses in a common cricket prey, Loxoblemmus equestris. We identified limonene as a behaviorally active volatile in bat odor that elicited electrophysiological responses in cricket antennae and was sufficient to elicit avoidance in crickets. Field experiments confirmed that limonene exposure reduced cricket calling activity, demonstrating the ecological relevance of this cue. Our findings establish that insects can detect and initiate avoidance of phylogenetically distant vertebrate predators via olfaction, a process that could be mediated by the elemental perception of individual odor compounds. This work broadens the sensory framework of a classic predator–prey system, and highlights olfactory eavesdropping as a functional strategy in phylogenetically distant predator–prey systems.

Introduction

The evolutionary arms race between insectivorous bats and their insect prey represents a quintessential predator–prey system, characterized by the evolution of sophisticated acoustic countermeasures in insects (Conner and Corcoran, 2012; Ter Hofstede and Ratcliffe, 2016). As dominant nocturnal aerial insectivores, bats consume diverse taxa such as Lepidoptera, Coleoptera, and Diptera (Gong et al., 2025; Srilopan et al., 2025), providing vital biocontrol against agricultural pests like Spodoptera frugiperda and Helicoverpa armigera (Liu et al., 2024; Maine and Boyles, 2015). Over 50 million years of coevolutionary arms race, insects evolved sophisticated auditory countermeasures: ultrasound hearing, sound-absorbing scales, acoustic decoys, and ultrasonic jamming (Barber et al., 2022; Jacobs and Bastian, 2016; Zeng et al., 2011). Auditory detection, independently evolved in at least seven insect orders, is the most widespread adaptation (Ter Hofstede and Ratcliffe, 2016). Upon detecting bat echolocation calls, insects typically execute evasive flight maneuvers away from the predator source to reduce detection and predation risk (Jacobs and Bastian, 2016). However, the dominance of this acoustic narrative may have overshadowed other sensory channels (Stidsholt et al., 2025).

Notably, both bats and insects themselves rely heavily on olfaction for essential behaviors. Many bats (e.g., Hipposideros armiger, Lasiurus cinereus, Phyllostomus hastatus, and Saccopteryx bilineata) produce species-specific volatile organic compounds (VOCs) from specialized glands for social communication (Adams et al., 2018; Caspers et al., 2009; Stefaniak et al., 2025; Zhang et al., 2022). Conversely, insects possess exquisitely sensitive olfactory systems capable of detecting chemical cues (e.g., sex pheromones) over kilometers, critical for foraging, mate finding, and predator avoidance (Kannan et al., 2022). Thus, olfaction could, in principle, provide an additional sensory channel in the bat–insect dynamic. More broadly, predator-derived VOCs are well-established as risk cues that elicit adaptive avoidance behaviors in diverse predator–prey systems (Apfelbach et al., 2005, 2015; Ferrero et al., 2011; Rosen et al., 2015; Takahashi et al., 2005, 2008). Empirical studies demonstrate that prey organisms exhibit multifaceted antipredator responses, including enhanced vigilance, activity suppression, and abandonment of non-defensive behaviors (e.g., foraging), when exposed to predator odors (Cornelis et al., 2019; Hegab et al., 2015; Pérez-Gómez et al., 2015). Such chemically cued defenses are well-established within phylogenetic groups: vertebrate prey (e.g., rabbits, rodents, and ungulates) avoid mammalian predator odors (e.g., bobcats, foxes, and wolves), while invertebrate prey (e.g., various arthropods including crustaceans and insects) evade arthropod predators (e.g., aquatic insect larvae and spiders) (Kempraj et al., 2020; Monclús et al., 2005; Osada et al., 2014; Swihart et al., 1991; Weiss et al., 2018). These intraphyletic frameworks elucidate core principles of chemical communication and inform biological control strategies (Apfelbach et al., 2005; Dicke and Grostal, 2001).

However, a fundamental gap exists in understanding whether such olfactory eavesdropping can operate across the vast phylogenetic divide separating vertebrate predators and invertebrate prey (Apfelbach et al., 2015; Dicke and Grostal, 2001; Schoeppner and Relyea, 2005). Although olfactory interactions across broad taxonomic boundaries are widespread in nature, such as mosquitoes using host odors to blood-feed, elephants and moths sharing pheromonal components, and aroids chemically mimicking carrion to attract pollinating flies (Kang et al., 2023; Zaremska et al., 2022; Zhao et al., 2022), these interactions are primarily shaped by selective pressures tied to foraging, reproduction, or mutualisms, rather than by predation-related selection. Given the pronounced differences between vertebrates and invertebrates in olfactory receptor architectures, neural processing, and behavioral ecology (Kaupp, 2010; Wang et al., 2024), two pivotal and interconnected questions arise: First, can invertebrates detect and respond to volatile cues from phylogenetically distant vertebrate predators? Second, what is the mechanistic basis of such recognition? Specifically, does it require the holistic perception of complex, multi-component odor blends (configural processing), or can it be triggered by the detection of simple, elemental compounds (Apfelbach et al., 2015; Clifford and Riffell, 2013; Coureaud et al., 2022; Lei and Vickers, 2008)? Resolving these questions is essential to advancing theories of sensory evolution and chemical ecology beyond intraphyletic paradigms.

The bat–insect system presents an ideal model to address these questions. Despite the clear importance of olfaction to both taxa and its established role in predator–prey ecology, whether it plays any functional role in the iconic bat–insect arms race remains unexplored. In the present study, we hypothesized that insects might eavesdrop on bat VOCs as an early warning. Using the Asiatic lesser yellow house bat (Scotophilus kuhlii) and its cricket prey (Loxoblemmus equestris) as a model system, we tested whether bat body odor elicits avoidance in insects and if individual odor components were sufficient to trigger such responses. S. kuhlii (Mammalia: Chiroptera: Vespertilionidae) is widely distributed across Asian tropics and subtropics. In China, this species typically forms dense colonies in palm trees (Livistona chinensis). S. kuhlii preys on a variety of insects, including members of the orders Coleoptera, Orthoptera, and Lepidoptera (Zhu et al., 2012). This species forages on the wing over forests, farmlands, and grasslands, habitats shared with L. equestris. This spatial overlap creates predation opportunities, as the bats frequently hunt near the ground. In our study area, L. equestris dominates cricket assemblages in the foraging habitats of S. kuhlii, producing intense mating calls (primarily for mate attraction and territory defense) and frequently relocating by flight, behaviors that heighten its vulnerability to predators (Ren et al., 2023; Stevenson and Rillich, 2012). Notably, S. kuhlii emits a strong and distinctive odor. Following confirmation of this specific predator–prey relationship by DNA metabarcoding, we integrated behavioral, electrophysiological, chemical, and field evidence to demonstrate olfactory-mediated avoidance in L. equestris and identify its underlying mechanism.

Results

L. equestris is a prey species of S. kuhlii

To establish the predatory relationship between S. kuhlii and L. equestris, we conducted DNA metabarcoding analysis on fecal samples from 30 individual bats. Using mitochondrial 16S rRNA gene (16S) and cytochrome c oxidase subunit I gene (COI) markers, we obtained 426,832 insect sequence reads (323 operational taxonomic units, OTUs) and 1,805,577 insect sequence reads (534 OTUs), respectively. The diet of S. kuhlii was primarily composed of Coleoptera, Blattodea, Lepidoptera, Hemiptera, Orthoptera, Mantodea, Diptera, and Hymenoptera (Figure 1A). Within Orthoptera, Gryllidae was the dominant family, comprising 94% (16S) and 93.5% (COI) of Orthoptera sequences (Figure 1B), and was detected in 50% of bats (15/30 individuals). Phylogenetic analysis revealed at least seven cricket species among the prey (Figure 1C). Among these prey crickets, L. equestris was the most abundant species encountered at ground level in the bat foraging sites. This ground-level abundance was supported by insect surveys conducted over eight nights during fecal sample collection, which captured 80 crickets, 70 (87.5%) of which were L. equestris (Supplementary Table 1). Based on the crickets’ well-developed olfactory capacity, the confirmed predation by S. kuhlii, the high local abundance and accessibility of L. equestris, and its tractability in behavioral settings, we selected L. equestris for subsequent experiments testing olfactory responses to bat body odor.

Dietary composition of Scotophilus kuhlii.

(A) Order-level composition based on 16S gene sequences, analyzed using the weighted percentage of occurrence (wPOO). Individual bats are labeled S1–S30. (B) Proportions of Orthoptera families in the diet, detected using the COI and 16S genes and presented as both relative read abundance (RRA) and weighted percentage of occurrence (wPOO). Color intensity corresponds to proportion magnitude. (C) Maximum-likelihood phylogenetic tree of Gryllidae reconstructed from 16S sequences. Operational taxonomic units (OTUs) identified in this study are highlighted in orange; reference sequences from NCBI are shown in black. Associated bat individuals are listed after each corresponding scientific name. Bootstrap support values are indicated at the nodes.

S. kuhlii body odor elicited behavioral avoidance and antennal responses in L. equestris

Behaviorally, in a two-choice olfactometer (Figure 2A), L. equestris strongly avoided the arm containing bat body odor. Upon release, 44 of 47 crickets (93.6%) chose the odor-free control arm, whereas only one entered the bat-odor arm and two shuttled without making a definitive choice (Chi-square test comparing avoiding vs. non-avoiding individuals: χ² = 35.76, df = 1, P < 0.001, φ = 0.87, 95% CI: 0.73 to 1.00; Figure 2B). Control trials with odor-free air in both arms showed no significant bias (13 of 24 vs. 11 of 24 crickets; Chi-square test, χ² = 0.17, df = 1, P = 0.68, φ = 0.08, 95% CI: 0 to 0.49; Figure 2B). Electrophysiological recordings using gas chromatography-electroantennographic detection (GC–EAD) confirmed that volatiles from S. kuhlii body odor are detected by L. equestris antennae. A mixture of bat odor volatiles from eight individuals elicited consistent antennal depolarizations (>0.1 mV) in all five crickets tested (Supplementary Figure 1). Parallel analysis by gas chromatography–mass spectrometry (GC–MS) identified six candidate compounds in the same odor samples. Among these, two compounds, namely 2,2-dimethylheptane and limonene, consistently co-eluted with EAD-active peaks (Figures 2C and 2D), identifying them as key antennal stimulants. No antennal responses were observed in odor-free controls tested on three crickets (Supplementary Figure 1). Together, these results establish that L. equestris perceives and avoids the body odor of its predator, S. kuhlii.

Behavioral and electrophysiological responses of L. equestris to S. kuhlii body odor.

(A) Schematic of the Y-tube olfactometer used in two-choice assays. The cricket’s release position (junction) and airflow direction are indicated. (B) Choice percentages of crickets for the control versus test arm across three treatments: bat body odor, air without bat odor (control for odor bias), and limonene. (C) Representative gas chromatography–electroantennographic detection (GC–EAD) recordings. Upper panel: flame ionization detection (FID) chromatograms of bat body odor extract (brown) and an odor-free control (blue). Lower panel: corresponding EAD response (black) of a L. equestris antenna. Peaks 1–6 correspond to compounds identified by GC–MS (see panel D); asterisks mark compounds that elicited consistent antennal depolarizations (>0.1 mV). (D) Gas chromatography–mass spectrometry (GC–MS) total ion chromatograms of the same bat body odor sample (upper) and an odor-free control (lower). Identified compounds are labeled (peaks 1–6). Axes indicate retention time (x) and relative ion intensity (y).

Snout secretions are the primary source of body odor in S. kuhlii

To identify the biological sources of bat body odor, we analyzed VOCs from hair, feces, and snout (pararhinal gland) secretions of nine bats using headspace solid–phase microextraction coupled with gas chromatography–mass spectrometry (HS–SPME–GC–MS). Snout secretions and hair shared similar hydrocarbon-rich VOC profiles, whereas fecal volatiles were distinct (Figure 3A). Individual-level terpenoid profiles further showed that limonene was detected in hair and snout secretion VOC collections but was absent from feces (Figure 3B and Supplementary Table 2). Principal component analysis revealed that the VOC profile of a pooled bat body odor sample (from eight bats) clustered more closely with those from snout secretions and hair than with feces (Figure 3C). As hair likely functions as a passive carrier, its surface VOCs probably originate from transferred snout secretions. These results point to snout secretions as the primary source of S. kuhlii body odor.

Volatile organic compound (VOC) profiles of potential odor sources in S. kuhlii.

(A) Hierarchical clustering of VOC profiles from feces, hair, and snout secretions, based on compositional similarity (values normalized 0–1). All three odor sources were sampled from the same nine bats (27 samples total). (B) Individual-level relative abundance of terpenoid compounds in VOC collections from hair, snout secretions, and feces. Each stacked bar represents one VOC collection from an individual bat, and the segment corresponding to limonene indicates its presence and relative contribution in the sample. (C) Principal component analysis (PCA) of VOC profiles from the same samples, based on compound presence/absence data, with the addition of a pooled body odor composite (combined from eight bats) for comparison. Ellipses represent 90% confidence intervals, and marginal density plots are shown for each source.

A single compound, limonene, triggers avoidance and reduces calling activity in L. equestris

To assess the behavioral relevance of identified volatiles, we measured antennal responses of L. equestris to four commercially available compounds previously detected in bat body odor: limonene, undecane, pentadecane, and hexadecane, using electroantennography (EAG). Compared to a hexane control, only limonene (10% v/v, 5.87 × 10−7 mol/µL) evoked a significant antennal response (repeated-measures ANOVA, F(4, 28) = 89.77, P < 0.001, η²p = 0.93; limonene vs. control: Bonferroni-corrected paired t-test, t(7) = 10.70, P < 0.001, Hedges’ g = 3.41; Figure 4A). EAG responses to limonene were concentration-dependent (repeated-measures ANOVA, F(5, 25) = 24.95, P < 0.001, η²p = 0.83; Figure 4B), with both 1% and 10% limonene solutions eliciting significantly stronger responses than the hexane control (Bonferroni-corrected paired t-tests, 1%: t(5) = 10.77, P < 0.001, Hedges’ g = 3.82; 10%: t(5) = 6.92, P = 0.005, Hedges’ g = 2.46).

Electrophysiological and ecological responses of L. equestris to candidate bat body odor compounds.

(A) Electroantennographic (EAG) responses to four candidate compounds (n = 8 individuals). Significance versus the hexane control is indicated (**P < 0.001; repeated-measures ANOVA with Bonferroni-corrected paired t-tests, see Results for exact F- and P-values and post-hoc comparisons). (B) Concentration-dependent EAG responses to limonene (n = 6 individuals). **P < 0.01, ***P < 0.001 versus control (same statistical test, see Results). (C) Mean call rate over time (calls per minute) in field plots treated with limonene (n = 8 datasets) or a hexane control (n = 7 datasets). Shaded areas represent 95% confidence intervals (1,200 recordings total).

Notably, limonene alone was sufficient to elicit robust avoidance behavior in L. equestris. In a two-choice olfactometer assay, 34 of 38 L. equestris individuals (89.5%) chose the limonene-free control arm, demonstrating significant avoidance (Chi-square test, χ² = 23.68, df = 1, P < 0.001, φ = 0.79, 95% CI: 0.59 to 0.99; Figure 2B). Field experiments confirmed the ecological relevance of this avoidance. Prior to spray application during their peak nocturnal activity period, the mean call rate of L. equestris did not differ significantly between treatment and control plots (mean ± SD, 245 ± 143 vs. 207 ± 133 calls/min; GLMM, estimate = 0.10, SE = 0.25, 95% CI: −0.39 to 0.59, Z = 0.38, P = 0.720; Supplementary Table 3). When exposed to limonene, the mean call rate of L. equestris in treatment plots significantly decreased over the 40-min trial (post- vs. pre-exposure: mean ± SD, 130 ± 139 vs. 245 ± 143 calls/min; GLMM, estimate = −0.68, SE = 0.06, 95% CI: −0.80 to −0.57, Z = −11.90, P < 0.001; Figure 4C and Supplementary Table 3). In contrast, call rates increased significantly in control plots sprayed with hexane alone (post- vs. pre-exposure: 298 ± 126 vs. 207 ± 133 calls/min; GLMM, estimate = 0.40, SE = 0.06, 95% CI: 0.28 to 0.51, Z = 6.68, P < 0.001; Figure 4C and Supplementary Table 3), a pattern consistent with their natural crepuscular activity peak. Together, these results demonstrate that limonene, a single component of bat odor, is sufficient to induce avoidance and reduce call activity in L. equestris under natural conditions.

Discussion

Our study provides direct experimental evidence that an insect species can detect and avoid a phylogenetically distant vertebrate predator via olfaction. We demonstrate that the body odor of the insectivorous bat S. kuhlii triggers robust avoidance and electrophysiological responses in the cricket L. equestris, and that a single compound, limonene, is sufficient to elicit this avoidance in the laboratory and suppress calling in the field. These findings fundamentally expand the sensory narrative of a classic predator–prey system and offer novel insights into the mechanisms of olfactory detection and chemical eavesdropping in interactions across broad phylogenetic divides.

Olfaction as a novel and functional channel in bat–insect interactions

For decades, the bat–insect arms race has served as a paradigm for acoustic coevolution, with insect auditory adaptations to echolocation being meticulously documented (Conner and Corcoran, 2012; Ter Hofstede and Ratcliffe, 2016). This dominant narrative, however, may have inadvertently narrowed the research focus, contributing to relatively less exploration of alternative sensory channels beyond audition (Stidsholt et al., 2025). Many bats produce body odor in the form of VOCs from specialized glands for intraspecific communication (Adams et al., 2018; Caspers et al., 2009; Stefaniak et al., 2025; Zhang et al., 2022). Here, we found that the body odor of S. kuhlii, which originates primarily from snout secretions, elicits robust avoidance and antennal responses in the cricket L. equestris. This suggests that insects may eavesdrop on predators’ intraspecific olfactory signals to mitigate predation threat. Our discovery thus reveals a previously overlooked olfactory dimension to this classic predator–prey interaction. This finding aligns with the established importance of multimodal predator detection, challenging the presumed supremacy of acoustics in this particular system (Jacobs and Bastian, 2016).

While our laboratory assays demonstrate clear olfactory-mediated avoidance in crickets, the ecological dynamics of a direct bat–insect encounter in the wild are more complex. Nonetheless, our field experiment provides critical, ecologically relevant support for the anti-predator function of this cue. The significant reduction in cricket calling activity upon limonene exposure—whether due to dispersal from the limonene-treated plots or direct cessation of signaling—constitutes a functionally adaptive anti-predator response, as both outcomes would reduce acoustic attraction and lower predation risk (Belwood and Morris, 1987; Hedrick and Kortet, 2006). This behavioral response confirms the ecological relevance of the olfactory cue, consistent with the evolutionary logic of eavesdropping on predator signals (Hermann and Thaler, 2014).

The functional advantages of this olfactory channel likely complement, rather than replace, established auditory defenses. The high sensitivity and long-range nature of insect olfaction may facilitate earlier predator detection, particularly against bats employing low-intensity or ‘stealth’ echolocation strategies that challenge acoustic detection alone (Corcoran and Conner, 2017; Goerlitz et al., 2010). More broadly, it could provide a vital, generalized defense for insect taxa lacking specialized ultrasonic hearing, including many beetles and non-hearing moths that are common bat prey (Göpfert and Hennig, 2016; Lin et al., 2023). The suppression of calling activity we observed exemplifies a key benefit: reducing acoustic conspicuousness. Insect calls can serve as a potential cue for certain bats, either through passive hearing or by enhancing the prey’s detectability within the returning echo (Alem et al., 2011; Falk et al., 2015; Prakash et al., 2021). We propose that olfaction in this system acts as a general alert mechanism, priming vigilance and enabling more informed deployment of evasive maneuvers triggered by closer-range acoustic (including echolocation) cues. In this integrated model, olfactory and auditory channels might operate across complementary spatial and temporal scales, forming a more robust, layered sensory defense system. Ultimately, direct observations of bat–insect encounters within manipulated olfactory environments will be crucial to fully quantify the survival benefit conferred by this eavesdropping channel under real predation pressure.

Implications for cross-phylum sensory ecology and the mechanism of eavesdropping

Beyond the specific bat–insect model, our work addresses a central question in sensory ecology: how chemical eavesdropping operates within predator–prey systems between phylogenetically distant taxa with fundamentally divergent olfactory systems (Adams et al., 2020; Emerson and Johnson, 2024; Kaupp, 2010). While intraphyletic kairomone detection is well-established (e.g., rodents avoiding carnivore odors, aphids responding to ladybug chemicals), compelling experimental evidence for such olfaction-mediated recognition across broad phylogenetic divides has been limited (Apfelbach et al., 2005; Ferrari et al., 2007; Tanis et al., 2018). By linking physiological detection to behavioral avoidance within a confirmed predator–prey relationship, we demonstrate that functional cross-phylum olfactory eavesdropping is attainable, proving this capacity persists despite profound divergence in underlying olfactory circuitry.

A key question concerns the perceptual mechanism underlying such olfactory eavesdropping: does it require a complex, species-specific odor blend (“configural processing”), or can it be initiated by simple, “elemental” compounds (Apfelbach et al., 2015; Coureaud et al., 2022; Lei and Vickers, 2008)? Our identification of limonene as a single, sufficient elicitor of avoidance supports an elemental perception strategy in this cross-phylum context. This mechanism is evolutionarily parsimonious. Evolving high-sensitivity receptors for a reliable kairomone bypasses the need to decode complex vertebrate odor profiles (Schoeppner and Relyea, 2005), likely being more efficient than evolving sophisticated neural circuits for configural processing (Haverkamp et al., 2018; Niven and Laughlin, 2008). Functionally, elemental perception enables a rapid, high-sensitivity response to trace amounts (Stengl, 2010), potentially allowing prey to identify threats at lower concentrations or greater distances than required for complex blends (Cardé and Willis, 2008; Minor and Kaissling, 2003). Thus, elemental perception emerges as a plausible and efficient strategy for initiating anti-predator behavior against phylogenetically remote threats.

Although limonene reliably induced avoidance behaviour in crickets, two related questions still merit careful consideration. One question is whether the limonene we identified genuinely originates from bats or reflects contamination during sampling. Limonene is common in plants and numerous consumer products (Boncan et al., 2020; Schuman, 2023), making its endogenous production by a mammal seem unusual. Nevertheless, multiple lines of evidence militate against contamination. First, we adhered to rigorous protocols. For example, all instruments were cleaned with ethanol and oven-dried before each use; bats were housed in stainless-steel cages, and cloth bags had been rinsed with purified water. Second, limonene was absent from all blank controls, including empty-chamber air samples and clean swabs, and was not detected in bat fecal samples. In contrast, it was consistently identified in hair and snout-secretion samples from bats. Third, independent studies have similarly identified limonene in the secretions of other bat species (Faulkes et al., 2019; Zhang et al., 2022). Furthermore, emerging evidence indicates skin-associated microbes may contribute to bat volatile profiles, with some taxa possessing enzymes involved in terpene biosynthesis (Sun et al., 2026). Taken together, these observations point towards an endogenous or microbe-mediated source, although the exact biosynthetic pathway remains to be determined.

The second question is how crickets might distinguish bat-derived limonene from environmental sources of this compound, given its prevalence in mint, citrus peel, pine and other non-threatening plants (Boncan et al., 2020; Schuman, 2023). It seems implausible that crickets could simply rely on limonene per se to differentiate a bat from a leaf. Two non-exclusive mechanisms could help resolve this issue. First, limonene need not be the only olfactory cue mediating risk perception. Our findings establish the sufficiency of limonene as an avoidance trigger, but do not preclude a role for other odor components. The crickets’ antennal responses to other bat volatiles in our GC–EAD analyses suggest more complex peripheral perception.

Additional compounds, either alone or in synergistic blends, may modulate the full behavioral response in nature. Therefore, elemental perception via limonene likely represents one effective strategy within a broader olfactory toolkit available to insects. Second, crickets may discriminate bat-derived limonene through context-specific cues (e.g., temporal and spatial patterning, co-occurrence with other bat-specific compounds) to minimize false alarms. Comparative studies on enantiomeric specificity and detection thresholds of cricket olfactory sensory neurons will be essential. Equally critical will be future efforts to quantify natural bat odor composition, limonene release rates, ambient exposure concentrations, and odor-plume dynamics, which together will inform ecologically valid stimulus design in controlled assays. Critically, our field data confirm that limonene exposure in nature robustly triggers an adaptive anti-predator response, irrespective of the precise discrimination mechanism.

In conclusion, our integrated approach reveals that the body odor of S. kuhlii triggers avoidance in L. equestris and that a single component, limonene, is sufficient to elicit this behavior in the laboratory and suppress calling in the field. We have uncovered a novel olfactory axis in the classic bat–insect arms race, demonstrating that chemical cues constitute a functional sensory channel for predator detection. More broadly, it provides evidence that olfactory eavesdropping can operate across vast phylogenetic distances within predator–prey systems. Elemental perception of individual odor compounds could represent a parsimonious strategy for such cross-phylum threat detection. Future work on the neural circuitry, phylogenetic distribution, and ecological contingencies of this capability will elucidate how such olfactory-mediated eavesdropping evolves and operates across the tree of life. The identification of behaviorally active volatiles like limonene may offer a starting point for the bioinspired development of novel insect repellents (Rasmann et al., 2005), but such applications must be informed by a deeper understanding of the ecological context in which these cues are perceived and interpreted.

Materials and methods

Diet composition and prey availability of Scotophilus kuhlii

To confirm the predator–prey relationship between S. kuhlii and L. equestris, we analyzed the diet of S. kuhlii via DNA metabarcoding of fecal samples. On July 10, 2022, we captured 30 individuals of S. kuhlii (21 females and nine males) from a roosting colony in a Chinese fan palm in Guangzhou, China, using hand nets. Sampling occurred during the bats’ initial return to the roost following nightly foraging activity, a period associated with high insect consumption. Each bat was placed individually in a thoroughly washed cotton bag, and fecal samples were collected after two hours. The samples were immediately transferred to 2-mL Eppendorf tubes (Corning, Cat# 430659) and stored at –20 °C. Bats were released unharmed after sample collection.

From each fecal sample, a 100-mg subsample was taken for DNA extraction using an Omega Bio-tek DNA extraction kit according to the manufacturer’s instructions. We targeted a 225-bp fragment of the mitochondrial cytochrome c oxidase subunit I (COI) gene with primers LCO-1490 (5′-GGTCAACAAATCATAAAGATATTGG-3′) and ZBJ-ArtR2c (5′-WACTAATCAATTWCCAAATCCTCC-3′), and a 105-bp fragment of the mitochondrial 16S ribosomal DNA with primers Coleop_16Sc (5′-TGCAAAGGTAGCATAATMATTAG-3′) and Coleop_16Sd (5′-TCCATAGGGTCTTCTCGTC-3′) (Lin et al., 2023). Polymerase chain reaction conditions followed published protocols (Lin et al., 2023). Amplified products of all 64 samples (60 from bat feces and four extraction blanks as negative controls) were sequenced on an Illumina MiSeq PE300 platform following the standard operating procedures at Majorbio Bio-Pharm Technology Co., Ltd. Ultimately, using the 16S molecular marker, we obtained insect sequences from 29 bats; no insect sequences were detected for one individual (bat identity: S28). Using the COI molecular marker, we obtained dietary sequences from 30 individual bats.

Dietary analysis was performed following established bioinformatic protocols (Chang et al., 2019; Lin et al., 2023). Briefly, after quality filtering and merging paired-end reads, DNA sequences were clustered into Operational Taxonomic Units (OTUs) at a 97% similarity threshold using Usearch v11.0 (Edgar, 2016). Taxonomic assignment was conducted by comparing representative sequences from each OTU against the GenBank and BOLD databases, based on the following criteria: (a) sequences with >98% identity to a single species were assigned to that species; (b) those matching multiple congeners within the same genus (>98%) were assigned to the genus level; and (c) sequences matching multiple genera within a family or only family-level references were assigned to the family level. No contaminant insect sequences were detected in the negative controls. Only insect sequences were retained for diet analysis. We quantified bat diet composition using the relative read abundance (RRA) and the weighted percentage of occurrence (wPOO) of prey orders, families and species (Deagle et al., 2018). For RRA, total read counts were normalized to 80,000 reads per sample. The wPOO was calculated from occurrence frequency tables standardized to a total occurrence frequency of 100 across all samples.

To achieve species-level resolution within Gryllidae (crickets), we implemented a phylogenetic approach (Ross et al., 2008). Representative sequences of each Gryllidae OTU were aligned with their top five BLAST matches from NCBI, and duplicate sequences were removed. Multiple sequence alignment was performed in MEGA v7.0 (Kumar et al., 2016), and a maximum-likelihood (ML) phylogeny was reconstructed using IQ-TREE v1.6.12 under the best-fit substitution model (16S: TPM3u+F+G4; COI: TIM2+F+I+G4) (Nguyen et al., 2015), with Gampsocleis sinensis sequences as the outgroup. OTUs clustering within established species-level clades were assigned to the corresponding species.

To assess prey availability, we surveyed insects in a grassland foraging site of S. kuhlii approximately one kilometer from its roost. Sampling was carried out on eight nights between July and September 2022 (19:30–21:30) using one ultraviolet light trap (BY201, Beijing Baoyuan Xingye Technology Co., Ltd., China) and one Malaise trap (BY201, Beijing Baoyuan Xingye Technology Co., Ltd., China). Lepidoptera (moths) and Gryllidae were the most abundant taxa collected. Preliminary Y-tube olfactometer trials indicated that moths exhibited minimal movement in the Y-tube, precluding reliable assessment of odor-mediated behavior, whereas crickets showed consistent locomotor activity. We therefore focused on estimating the relative abundance of Gryllidae species. L. equestris was selected for subsequent olfactory experiments due to its high local abundance and confirmed status as a natural prey item of S. kuhlii (see Results).

Behavioral and electrophysiological assays of cricket responses to S. kuhlii body odor

We employed two-choice tests in a Y-tube olfactometer (Shelai Instrument Platform, China) to assess whether the body odor of S. kuhlii elicits avoidance behavior in the cricket L. equestris. Individuals of L. equestris were collected nightly (19:30–21:30) from the bat’s foraging habitat (see the section above) and acclimated in insect cages with water provided in a temporary laboratory until testing.

The olfactometer consisted of a Y-shaped glass tube (inner diameter: 35 mm; main arm length: 200 mm; side arm length: 200 mm). One side arm was connected in series to a glass chamber housing one of eight adult S. kuhlii individuals (randomly selected for each trial) and a charcoal filter bottle; the other side arm was connected to an empty glass chamber and a charcoal filter bottle. Both charcoal bottles contained activated carbon for air purification. An air pump supplied airflow, regulated to 1 L/min in each arm by flowmeters.

Behavioral trials commenced at 23:00 under complete darkness. Before each trial, the Y-tube was cleaned with ethanol and oven-dried. Odorless sponges (diameter: 35 mm, thickness: 15 mm; Wuhan Hemida Technology Co., Ltd., China) were placed at the end of each arm to prevent the crickets from escaping. After confirming system airtightness, a single cricket was introduced at the junction (center) of the Y-tube, and its behavior was recorded using an infrared video camera (4K Ultra; Sony, China). Preliminary observations confirmed that crickets typically initiated exploration of the apparatus arms within three seconds of being introduced. Therefore, video recording began once a cricket started exploring the two side arms. A cricket was scored as “avoiding” the bat odor if it entered the control arm and remained there continuously for 10 s. It was scored as “not avoiding” the bat odor if it entered the test arm and remained there continuously for 10 s. Individuals that moved back and forth between the two arms without remaining in either arm for at least 10 s were also classified as “not avoiding”. We selected a 10-second duration as the criterion because the crickets were highly mobile, and this period was sufficient for them to fully explore the olfactometer. A total of 47 crickets were tested, with each individual used in a single trial. The left/right position of the bat odor source was randomized between trials to control for side bias. A separate control experiment with 24 crickets was performed under identical conditions, except that no bat was placed in the glass chamber. Behavioral responses were recorded following the same procedure. To minimize observer bias, the video recordings were analyzed by an experimenter who was blinded to the treatment conditions.

We collected the body odor of S. kuhlii (i.e., its volatile organic compounds, VOCs) for chemical characterization and electrophysiological bioassay using gas chromatography–mass spectrometry (GC–MS) and gas chromatograph–electroantennographic detection (GC–EAD) with cricket antennae, respectively. Samples were collected from eight adult individuals (four females and four males) using a dynamic headspace sampling system equipped with Porapak-Q adsorbent columns. We selected Porapak-Q because bat body odor is rich in hydrocarbons, which are non-polar or only weakly polar, and Porapak-Q has a high adsorption capacity for these compounds. The system consisted of a vacuum pump (QC-1S; Beijing Institute of Labour Protection Science, China), a charcoal filter for air purification, a 5 L glass chamber housing the bats, and a Porapak-Q column (glass tube, 4 mm internal diameter, 200 mg of Porapak-Q 80/100 mesh; Supelco, USA) (Wang et al., 2023). Prior to sampling, all glassware was thoroughly rinsed with ethanol and dried in an oven at 120°C, and volatile odor collection was conducted in a dedicated odor-free room to minimize environmental contamination. Each adsorbent column was cleaned by rinsing sequentially with 3 mL of methanol and dichloromethane (DCM), followed by thermal conditioning under nitrogen at 180 °C for 30 minutes. During VOC collection, purified air from the pump was passed through the system at a flow rate of 500 mL/min for two hours. Control samples were collected from empty glass chambers following the same procedure. After sampling, VOCs were eluted from each column with 1 mL of DCM, and the extracts were concentrated under a gentle nitrogen stream to a final volume of 1 mL. All samples were stored at –20 °C until subsequent analysis by GC–EAD and GC–MS. The eight S. kuhlii used for odor collection and behavioral assays were captured from a roosting palm tree on 1 October, 2023. Bats were housed socially in cages (50 × 50 × 50 cm) in a temporary laboratory maintained at 28 °C and 60% relative humidity. Mealworm larvae (Tenebrio molitor) and water were provided daily following nightly experiments.

The antennae are a primary sensory organ for detecting both chemical and mechanical cues in crickets. To determine if L. equestris can physiologically perceive bat body odor, we employed a GC–EAD system following established methods (Barbosa-Cornelio et al., 2019; Ma et al., 2025) to test the electrophysiological sensitivity of L. equestris antennae to bat-derived VOCs. The system consisted of an Agilent 7820A gas chromatograph (Agilent Technologies, USA) and EAD equipment (Syntech, Germany). The EAD setup comprised an IDAC 4 signal recording controller, an MP-15 micromanipulator with high-impedance PRG-3 electrodes.

The tips and bases of cricket antennae were trimmed and mounted on a PRG-3 electrode using an MP-15 micromanipulator. After achieving a stable baseline, odor samples were separated on an Agilent 7820A gas chromatograph. A 4 μL aliquot was injected in splitless mode onto an HP-5 column (30 m × 0.32 mm × 0.25 μm). We selected the HP!Il5 column because its non!Ilpolar stationary phase achieves excellent chromatographic separation of the non!Ilpolar to weakly polar hydrocarbons that dominate bat body odor, which is essential for subsequent detection. The GC oven temperature was programmed as follows: held at 50 °C for 5 min, increased at 10 °C/min to 200 °C (held 1 min), then raised at 20 °C/min to 250 °C (held 5 min). Nitrogen carrier gas flow was 2.0 mL/min. Injector and detector temperatures were 250 °C and 300 °C, respectively. The GC effluent was split at a 1:2 ratio using a Y-shaped splitter. Two-thirds of the effluent was directed to a flame ionization detector (FID) for compound detection, and one-third was transferred through a heated transfer line into a humidified airstream (400 mL/min) before being delivered to the cricket antenna. Antennal signals were acquired using an IDAC-4 interface and GcEad software (v. 1.2.5; Syntech, Germany), allowing simultaneous recording of FID and EAD responses. Each chromatographic peak in the FID trace had a corresponding EAD trace showing the depolarization of the antennal potential.

A total of eight antennae from individual crickets were tested. Five antennae were exposed to bat odor extracts, and three were exposed to empty-chamber controls (one antenna per injection). The empty-chamber controls were used to account for potential background signals from the experimental system and to provide a baseline for comparison with bat odor extracts. Snout secretions were not included in the GC–EAD analyses because the present study focused on bat whole-body odor as the ecologically relevant airborne odor source, and sufficient secretion samples for standardized volatile extraction were not available for electrophysiological testing. A compound was considered EAD-active if its GC peak coincided with an antennal depolarization >0.1 mV, and this response was reproducible in at least three individual antennae. Active compounds were identified by aligning the retention times and peak profiles of EAD-active signals with those from parallel GC–MS analyses.

Chemical profiling and source analysis of S. kuhlii body odor

To characterize the major volatile constituents of S. kuhlii body odor, we analyzed the collected VOC extracts by GC–MS using an Agilent 6850-5975 mass spectrometer operated in electron ionization (EI) mode at 70 eV. The GC was equipped with the same HP-5 column used for GC–EAD, with helium as the carrier gas at a constant flow of 1 mL/min. A 1 μL aliquot of each sample was injected in splitless mode (0.75-min splitless period) at an injector temperature of 250 °C.

The oven temperature program matched that used for GC–EAD. Compounds were tentatively identified by comparison of their mass spectra with the NIST 2017 library (Supplementary Table 4; Supplementary Figure 2–7). The identities of these compounds were confirmed by matching both retention times and mass spectra to those of authentic standards, including injections of a mixture containing 100 ng of each compound. However, co-injection with authentic standards could not be performed because the total extract volume from each sample was insufficient for additional runs after the primary GC-MS and GC-EAD analyses, owing to the limited amount of biological material available.

To investigate potential biological sources and characterize the volatile chemical profiles of bat body odor, we performed a broadly targeted volatile metabolomic analysis using headspace solid–phase microextraction coupled with GC–MS (HS–SPME–GC–MS) on hair, pararhinal gland (snout) secretions, and feces from nine bats (four females, five males). These bats were collected and maintained as described above. Upon capture, bats were placed in clean stainless-steel cages and kept in groups consistent with their natural social associations during the brief interval prior to immediate odor sampling. Hair samples (10 mg per individual) were clipped from dorsal and ventral regions. Snout secretions were collected using sterile cotton swabs (CS15-005, Shenzhen SihuaBo Technology Co., Ltd., China), with two blank swabs as controls. These blank swab controls were included to account for potential volatile contamination from ambient air or the swab material itself (Supplementary Table 5). Fecal samples (100 mg per individual) were collected as in the dietary analysis. All samples were stored in 2-ml Eppendorf tubes at −80 °C until analysis.

Samples were thawed on ice, transferred to headspace vials, and spiked with 10 µL of an internal standard (50 µg/mL). Volatiles were extracted by HS–SPME: after incubation at 60 °C for 5 min, a 120 μm DVB/CWR/PDMS fiber was exposed to the headspace for 15 min for adsorption, then thermally desorbed at 250 °C for 5 min in the GC injector. Analysis was performed on an Agilent 7890B 7000D GC–MS system equipped with a DB 5MS capillary column (30 m × 0.25 mm × 0.25 μm), with helium at 1.2 mL/min. The oven temperature program was: 40 °C for 3.5 min, then increased at 10 °C/min to 100 °C, 7 °C/min to 180 °C, and 25 °C/min to 280 °C (hold 5 min). The MS operated in EI mode at 70 eV, with ion source, quadrupole, and interface temperatures set at 230 °C, 150 °C, and 280 °C, respectively. Data were acquired in selected ion monitoring (SIM) mode. Each compound was identified by matching the retention time and the presence of one quantitative and two to three qualitative ions to those of a standard reference (Yuan et al., 2022). Data were integrated, corrected, and processed using MassHunter software (B.08.00; Agilent), and metabolites were qualified and quantified against an in-house database (Supplementary Table 5). Retention times for non-target compounds are not fully disclosed due to the proprietary policy of the commercial analytical service provider.

To identify the biological source of the characteristic body odor, we compared the VOC profiles from hair, feces, and snout secretions. Volatiles were classified into mutually exclusive chemical classes for comparison; note that “Hydrocarbons” here refers only to aliphatic hydrocarbons, and “Nitrogen compounds” excludes amines. To objectively assess the overall compositional similarity among hair, feces, snout secretions, and body odor, we performed a principal component analysis (PCA) based on binary (presence/absence) VOC data.

Identifying avoidance-eliciting odor compounds

Chemical profiling of S. kuhlii body odor by GC–MS identified six volatile compounds: 2,2-dimethylheptane, limonene, undecane, 2,6,7-trimethyldecane, pentadecane, and hexadecane, two of which elicited EAD responses (Figure 2C). Because GC–EAD provides a physiological readout of antennal sensitivity but may not fully capture behavioral relevance across different concentrations or odor contexts, we additionally tested antennal responses to commercially available standards of compounds identified by GC–MS. We selected four compounds (limonene, undecane, pentadecane, and hexadecane) and evaluated antennal responses using electroantennography (EAG) (Supplementary Table 6). Each compound was diluted to 10% (v/v) in hexane for screening, resulting in a final limonene concentration of 5.87 × 10−7 mol/µL. Hexane alone served as the solvent control. Stimuli were prepared by applying 10 μL of a test solution to a filter-paper strip (10 mm × 30 mm), which was then placed inside a glass Pasteur pipette. The pipette tip was positioned in a humidified airstream (1500 mL/min) directed at the antenna through a stainless-steel tube (7 mm inner diameter, 1 cm from the antenna). Cricket antennae were prepared as for GC–EAD. For each antenna, a hexane control was first presented to establish baseline antennal activity. For the initial screening, limonene, undecane, pentadecane, and hexadecane were diluted to 10% (v/v) in hexane and delivered individually in a randomized order. To assess dose-dependent responses, limonene was further tested at five concentrations (0.001%, 0.01%, 0.1%, 1%, and 10%, v/v in n-hexane), following the concentration gradient used in a previous study (Tang et al., 2024). Following the initial hexane control, the five limonene concentrations were tested in a randomized order across trials. Each stimulus lasted 0.5 s, with an inter-stimulus interval of 1 min to allow full recovery of antennal responses. EAG responses were recorded from eight antennae for the initial screening and from six antennae for the dose–response assay, with each antenna obtained from a different individual, and analyzed using EagPro software (version 2.0; Syntech, Germany).

To examine the behavioral response of L. equestris to limonene, a two-choice assay was conducted using 10% (v/v) limonene (5.87 × 10−7 mol/µL) as the odor source. A 10 μL aliquot was applied to a filter-paper strip (10 mm × 30 mm) inside a glass chamber. A total of 38 L. equestris individuals were tested under the same experimental conditions as in the bat-odor assay.

To ecologically validate that a single odor compound can elicit avoidance, we tested the effect of 10% (v/v) limonene (5.87 × 10−7 mol/µL) on the calling activity of L. equestris within the bats’ foraging habitat. Over nine nights between 24 June to 16 July 2024, we established four sampling plots (5 m × 5 m) in a linear array at 100-m intervals within the known foraging range of S. kuhlii, at the same location where crickets were originally captured (Supplementary Figure 8). Each night, two plots were randomly assigned to the limonene treatment and two to hexane control. To assess avoidance, we monitored changes in calling activity of L. equestris as a proxy, using passive acoustic monitoring during their peak nocturnal activity. At each plot, an Audiomoth (v 1.2.0, Open Acoustic Devices, UK) was positioned 0.5 m above ground with the microphone facing downward. Audio was recorded in 55-s files separated by 5-s intervals (1-min duty cycle) at a 384 kHz sampling rate (medium gain). A 40-min pre-exposure recording started at 19:50. Between 20:30 and 20:40, 1 mL of 10% limonene (treatment) or hexane (control) was evenly sprayed over the 25 m2 area centered on the recorder. A 40-min post-exposure recording started at 20:50. Experimental nights were separated by at least one day, and rainy conditions were avoided.

During acoustic preprocessing, we excluded plot–night datasets that failed quality-control criteria: no cricket calls detected, >20% of 1-min files with zero calls, or total call count <10,000. This quality-control procedure yielded a final sample of 15 datasets: 8 from the limonene-treated (experimental) plots and 7 from the hexane-treated (control) plots. In total, 1,200 recordings (80 files/plot/night) were analyzed. Using Avisoft SASLab Pro (v 5.2.07, Avisoft Bioacoustics, Germany), recordings were band-pass filtered (3–6.5 kHz) to match the calling frequency of L. equestris (dominant frequency ∼5 kHz, range 4.5–5.5 kHz). This frequency band does not overlap with sympatric cricket species, and L. equestris is the dominant local cricket (see Results), allowing reliable identification. Calls were identified via spectrogram matching using 10 templates of L. equestris calls. A “call” was defined as a train of successive sound elements (syllables). Typical calls of L. equestris last 60–200 ms (Supplementary Figure 9), with inter-syllable intervals < 10 ms within a call, and >60 ms between consecutive calls from the same individual. The number of calls per minute was quantified for each plot and recording period. Our assays were designed to test physiological detectability and functional sufficiency, rather than to establish concentration thresholds or mimic natural emissions exactly.

Data analysis

All analyses were performed in R v4.3.2 unless specified otherwise. For the two-choice olfactory assay, we used a chi-square goodness-of-fit test to compare the number of crickets exhibiting “avoiding” versus “not avoiding” behavior. Effect sizes are reported as the phi coefficient (φ) with 95% confidence intervals (estimated via the delta method). For chemical profiling, we performed multivariate analysis on VOC data. Hierarchical clustering with heatmap visualization was conducted in Python (v3.9) using the pandas, seaborn, and matplotlib libraries, based on data normalized to a 0–1 range, Euclidean distance, and average linkage. Principal component analysis (PCA) based on a binary (presence/absence) matrix was performed using the vegan package in R to compare profiles from hair, feces, snout secretions, and bat body odor. For the EAG experiment, EAG responses were analyzed using repeated-measures ANOVA (sphericity checked with the rstatix package), followed by Bonferroni-corrected paired t-tests. Effect sizes are reported as partial eta-squared (η²p) for the ANOVA and Hedges’ g for the t-tests. For field calling activity, a generalized linear mixed model (GLMM) was fitted to calls per minute using the glmmTMB package. Model selection compared zero-inflated and standard negative binomial distributions using AIC, and evaluated alternative random effects structures and fixed effects using AICc. The final model included a Group × Phase interaction as a fixed effect, with a random intercept for the plot-by-date combination to account for spatial and temporal non-independence. Model diagnostics were performed with the DHARMa package, including tests for residual uniformity, overdispersion, and zero-inflation.

Animal welfare and ethics

The bats used in this study were carefully handled and maintained under controlled conditions (28 °C, 60% relative humidity) with ad libitum access to food and water. Bats were housed socially in cages (50 cm × 50 cm × 50 cm). No bat mortality occurred during the study, and all bats were released at their original roost sites upon completion of experiments. Crickets were kept at densities ≤ 5 individuals per container to minimize aggression. Crickets not used in EAG assays were returned to their collection sites. All procedures involving live animals were approved by the Science and Technology Ethics Committee of Northeast Normal University, Changchun, China (permit ID: NENU-2022-0308).